---
title: "EcoTox_MonteCarlo"
output: html_document
date: "2024-07-19"
output:
  html_document:
    code_folding: hide
    theme: journal
    toc: yes
    toc_float: yes
    toc_depth: 6
    number_sections: true
  word_document:
    toc: yes
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Setup
## Libraries
```{r}
library(tidyverse)
library(sensobol)
library(truncnorm)
library(ggpubr)
library(ggdark)
library(gtsummary)
```

## Functions
```{r}
getwd()
set.seed(123456789) 
# each script is now a function. Source files to load functions #
source("RDAmaker_functions.R") #functionalized data tidying script for ToMEx 1.0 data
source("ToMEx2.0_Data_Tidying_functions.R") #functionalized data tidying script for ToMEx 2.0 data
source("ssd_functions.R")
source("../../scripts/functions.R")
```

## Parameters and Distributions
### Static Parameters
```{r}
## Monte Carlo Params ##
nboot = 10 #ssd bootstraps
n_sim <- 10 #monte carlo simulations

## Alingment Params ##
x1M_set <- 1 #um lower size for all alignments
x1D_set <- 1 #um lower size for all alignments
x2D_set <- 5000 #um

upper.tissue.trans.size.um <- 83 #10 #um #set size for x2M #only used for determinstic assessment
```

### Dynamic Parameters
```{r}
## Distribution values
# particle properties
R.ave.water.marine <- 0.77 # average length to width ratio of microplastics in marine environment (Kooi et al. 2021)
R_ave.water.marine <- 0.77 # average length to width ratio of microplastics in marine environment (Kooi et al. 2021)
R.ave.water.marine.sd = 0.29 #Tablse s3. Marine surface water
R.ave.water.marine_samples <- rnorm(n_sim, mean = R.ave.water.marine, sd = R.ave.water.marine.sd)

R.ave.water.freshwater <- 0.67 # average length to width ratio of microplastics in freshwater environment (Kooi et al. 2021)
R_ave.water.freshwater <- 0.67 # average length to width ratio of microplastics in freshwater environment (Kooi et al. 2021)
R.ave.water.freshwater.sd = 0.28 #Tablse s3. freshwater surface water
R.ave.water.freshwater_samples <- rnorm(n_sim, mean = R.ave.water.freshwater, sd = R.ave.water.freshwater.sd)

R.ave.sediment.marine <- 0.75 # average length to width ratio of microplastics in marine environment (Kooi et al. 2021)
R.ave.sediment.marine.sd = 0.30 #Tablse s3. Marine surface sediment
R.ave.sediment.marine_samples <- rnorm(n_sim, mean = R.ave.sediment.marine, sd = R.ave.sediment.marine.sd)

R.ave.sediment.freshwater <- 0.70 # average length to width ratio of microplastics in freshwater environment (Kooi et al. 2021)
R.ave.sediment.freshwater.sd = 0.33 #Tablse s3. freshwater surface sediment
R.ave.sediment.freshwater_samples <- rnorm(n_sim, mean = R.ave.sediment.freshwater, sd = R.ave.sediment.freshwater.sd)

#### MARINE ####
p.ave.marine = 1.10 #average density in marine surface water
p.ave.sd.marine = 0.14 #Tablse s3. Marine surfac water
p.ave.marine_samples <- rnorm(n_sim, mean = p.ave.marine, sd = p.ave.sd.marine)

# Alignment properties
alpha.marine = 2.07 #table s4 for marine surface water. length
alpha_marine = 2.07 #table s4 for marine surface water. length
alpha.sd.marine = 0.03 #table s4 for marine surface water. lengthj
alpha.marine_samples <- rnorm(n_sim, mean = alpha.marine, sd = alpha.sd.marine)

# define parameters for power law coefficients
a.sa.marine = 1.50 #marine surface area power law
a_sa.marine = 1.50 #marine surface area power law
a.sa.sd.marine = 0.009 #marine surface water surface area power law - table s4
a.sa.marine_samples <- rnorm(n_sim, mean = a.sa.marine, sd = a.sa.sd.marine)

a.v.marine = 1.48 #a_V for marine surface water volume
a_v.marine = 1.48 #a_V for marine surface water volume
a.v.sd.marine = 0.063
a.v.marine_samples <- rnorm(n_sim, mean = a.v.marine, sd = a.v.sd.marine)

a.m.marine = 1.32 # upper limit fora_m for mass for marine surface water in table S4
a_m.marine = 1.32 # upper limit fora_m for mass for marine surface water in table S4 
a.m.sd.marine = 0.009
a.m.marine_samples <- rnorm(n_sim, mean = a.m.marine, sd = a.m.sd.marine)

a.ssa.marine = 1.98 # A_SSA for marine surface water
a_ssa.marine = 1.98 # A_SSA for marine surface water
a.ssa.sd.marine = 0.297
a.ssa.marine_samples <- rnorm(n_sim, mean = a.ssa.marine, sd = a.ssa.sd.marine)

#### FRESHWATER ####
p.ave.freshwater = 1.04 #average density in freshwater surface water
p.ave.sd.freshwater = 0.12 #Tablse s3. freshwater surfac water
p.ave.freshwater_samples <- rnorm(n_sim, mean = p.ave.freshwater, sd = p.ave.sd.freshwater)

# Alignment properties
alpha.freshwater = 2.64 #table s4 for freshwater surface water. length
alpha_freshwater = 2.64 #table s4 for freshwater surface water. length
alpha.sd.freshwater = 0.01 #table s4 for freshwater surface water. lengthj
alpha.freshwater_samples <- rnorm(n_sim, mean = alpha.freshwater, sd = alpha.sd.freshwater)

# define parameters for power law coefficients
a.sa.freshwater = 2.00 #freshwater surface area power law
a_sa.freshwater = 2.00 #freshwater surface area power law
a.sa.sd.freshwater = 0.065 #freshwater surface water surface area power law - table s4
a.sa.freshwater_samples <- rnorm(n_sim, mean = a.sa.freshwater, sd = a.sa.sd.freshwater)

a.v.freshwater = 1.68 #a_V for freshwater surface water volume
a_v.freshwater = 1.68 #a_V for freshwater surface water volume
a.v.sd.freshwater = 0.081
a.v.freshwater_samples <- rnorm(n_sim, mean = a.v.freshwater, sd = a.v.sd.freshwater)

a.m.freshwater = 1.65 # upper limit fora_m for mass for freshwater surface water in table S4
a_m.freshwater = 1.65 # upper limit fora_m for mass for freshwater surface water in table S4 
a.m.sd.freshwater = 0.071
a.m.freshwater_samples <- rnorm(n_sim, mean = a.m.freshwater, sd = a.m.sd.freshwater)

a.ssa.freshwater = 2.71 # A_SSA for freshwater surface water
a_ssa.freshwater = 2.71 # A_SSA for freshwater surface water
a.ssa.sd.freshwater = 0.009
a.ssa.freshwater_samples <- rnorm(n_sim, mean = a.ssa.freshwater, sd = a.ssa.sd.freshwater)

# Coefficients
beta_log10_body_length <- 0.9341
body_length_intercept <- 1.1200

# Standard errors for these coefficients
se_beta_log10_body_length <- 0.1376  # SE generated from Jams et al R code (not rpoerted in paper or SI!) (https://github.com/fmwindsor/plastic-allometry)
se_body_length_intercept <- 0.3222      # SE generated from Jams et al R code (not rpoerted in paper or SI!) (https://github.com/fmwindsor/plastic-allometry)
#data
sim_beta_log10_body_length_samples <- rnorm(n_sim, mean = beta_log10_body_length, sd = se_beta_log10_body_length)
sim_body_length_intercept_samples <- rnorm(n_sim, mean = body_length_intercept, sd = se_body_length_intercept)

#### Tissue translocation length ##
# Parameters from the logistic regression# See scripts/translocation/translocation.Rmd for model
beta_0 <- 1.308344# simple$coefficients[[1]]  # 1.24  # Intercept
beta_1 <- -0.01468148 # simple$coefficients[[2]]   #-0.014  # Slope for particle length
se_beta_0 <- 0.3963612 # simple_summary$coefficients[[1,2]]  # 0.40  # SE of intercept
se_beta_1 <- 0.006657993 #simple_summary$coefficients[[2,2]]  # SE of slope

# Simulate beta_0 and beta_1
sim_beta_0 <- rnorm(n_sim * 1.2, mean = beta_0, sd = se_beta_0)
sim_beta_1 <- rnorm(n_sim * 1.2, mean = beta_1, sd = se_beta_1)

# Calculate X50 for each simulation
sim_X50 <- -sim_beta_0 / sim_beta_1

#truncate distribution to not fall below 0
upper.tissue.trans.size.um_samples <- sim_X50 %>% data.frame() %>% filter(.>0) %>% slice(1:n_sim)
upper.tissue.trans.size.um_samples <- as.numeric(upper.tissue.trans.size.um_samples$.)

#define param values
param_values <- data.frame(
  alpha.marine = alpha.marine_samples,
  a.sa.marine = a.sa.marine_samples,
  a.v.marine = a.v.marine_samples,
  a.m.marine = a.m.marine_samples,
  a.ssa.marine = a.ssa.marine_samples,
  R.ave.water.marine = R.ave.water.marine_samples,
  alpha.freshwater = alpha.freshwater_samples,
  a.sa.freshwater = a.sa.freshwater_samples,
  a.v.freshwater = a.v.freshwater_samples,
  a.m.freshwater = a.m.freshwater_samples,
  a.ssa.freshwater = a.ssa.freshwater_samples,
  R.ave.water.freshwater = R.ave.water.freshwater_samples,
  #R.ave.sediment.marine = R.ave.sediment.marine_samples,
  #R.ave.sediment.freshwater = R.ave.sediment.freshwater_samples,
  sim_beta_log10_body_length = sim_beta_log10_body_length_samples,
  sim_body_length_intercept = sim_body_length_intercept_samples,
  upper.tissue.trans.size.um =  upper.tissue.trans.size.um_samples
)
```

### Visualize Distribution Parameters
```{r}
MC_param_histograms <- param_values %>% 
  pivot_longer(cols = everything()) %>% 
  mutate(name = factor(name, levels = c("alpha.marine", "a.sa.marine", "a.v.marine",
                                        "a.m.marine", "a.ssa.marine", "R.ave.water.marine",
                                        "alpha.freshwater", "a.sa.freshwater", "a.v.freshwater",
                                        "a.m.freshwater", "a.ssa.freshwater", "R.ave.water.freshwater",
                                  "sim_beta_log10_body_length","sim_body_length_intercept","upper.tissue.trans.size.um"))) %>% 
  ggplot(aes(x = value, fill = name)) +
  geom_histogram(bins = 350, linewidth = 0.10) +
  scale_x_log10()+
#  xlim(c(1,1000))+
  labs(#title = "Distribution of Simulated X50 Values",
       x = "Value",
       y = "Relative Frequency") +
  theme_minimal(base_size = 12) +
  theme(legend.position = "none",
        plot.title = element_blank(),
        plot.subtitle = element_text(hjust = 0.5),
        plot.caption = element_text(hjust = 0.5),   # Center the caption
        axis.title.x = element_text(hjust = 0.5, size = 16, face = "bold"),  # Center the x-axis title
        axis.text.x = element_text(size = 11),
        axis.text.y = element_blank(),
       # axis.text = element_text(size = 16),
        axis.title.y = element_text(vjust = 0.5, size = 16, face = "bold"),
       strip.text = element_text(size = 16, face = "bold", family = "serif")  # Customize facet title size
       )  + # Center the y-axis title
  facet_wrap(~name, scales = "free",
             ncol = 6)

ggsave(filename = "MC_param_histograms.jpg",
       dpi = 300,
       path = "../output/Manuscript_Figs/", 
       plot = MC_param_histograms, width = 9, height = 8, units = "in")

MC_param_histograms
```
### Summary Statistics of Distributions
```{r}
skimr::skim(param_values)
```


### Export Dataframe of distribution parameters
```{r}
write.csv(param_values, "output/parameter_values.csv")
param_values
```


## Generate Base dataframes
```{r warning=FALSE}
#### read in input data necessary for functionalized data tidying 
# ToMEx 1.0 data file (no alignments)
aoc <- read_csv("ref data/AquaticOrganisms_Clean_final.csv", guess_max = 10000, show_col_types = FALSE) %>% rowid_to_column()
# ToMEx 2.0 data file (no alignments)
tomex2.0 <- read_rds("ref data/tomex2_input.rds")
# auxiliary data necessary for ToMEX 2.0 data tidying
bodysize_addons <- read_csv("ref data/gape_size.csv",show_col_types = FALSE) #copied from aq_mp_tox_shiny main folder

# these functions perform alignments
aoc_setup <- ToMEx1.0fxn(R.ave.water.marine = R.ave.water.marine,
                         R.ave.water.freshwater = R.ave.water.freshwater,
                         R.ave.sediment.marine = R.ave.sediment.marine,
                         R.ave.sediment.freshwater = R.ave.sediment.freshwater,
                         beta_log10_body_length = beta_log10_body_length,
                         body_length_intercept = body_length_intercept) 

## generate ToMEx 2.0 dataset ###
tomex2.0_aoc_z_final <- ToMEx2.0fxn(aoc_setup = aoc_setup,
                                    R.ave.water.marine = R.ave.water.marine,
                                    R.ave.water.freshwater = R.ave.water.freshwater,
                                    R.ave.sediment.marine = R.ave.sediment.marine,
                                    R.ave.sediment.freshwater = R.ave.sediment.freshwater,
                                    beta_log10_body_length = beta_log10_body_length,
                                    body_length_intercept = body_length_intercept)
```


# Monte Carlo
## Functions
### Alignment functions
```{r}
alignmentFxn <- function(df = tomex2.0_aoc_z_final){
  
 ### perform alignments
  aoc_MC_iter <- df  %>% 
    ungroup() %>% 
    rename(environment = env_f) %>% 
    ### First filter the data ####
  ## First filter data with global filters
  filter(!environment %in% c("Terrestrial", "Not Reported"),
         Group != "Bacterium",
         Group != "Plant",
         effect.metric != "HONEC",
         tier_zero_tech_f == "Red Criteria Passed",
         tier_zero_risk_f == "Red Criteria Passed", #All thresholds must pass technical and risk red criteria
         risk.13 != 0 #Drop studies that received a score of 0 for endpoints criteria (this also drops studies that have not yet been scored) - KEEP THIS AFTER THE RED CRITERIA FILTERS  
  ) %>% 
    #Remove 26C temperature treatment data from Jaimukar et al. 2018
    filter(!(article == 42 & media.temp == 26)) %>% 
    mutate(max.size.ingest.um = 1000 * max.size.ingest.mm) %>%  #makes it less confusing below
    # calculate ERM for each species
    #### TISSUE TRANSLOCATION ####
  # define upper size length for Translocation 
  #set to 83um for upper limit or max size ingest, whichever is smaller
  mutate(x2M_trans = case_when(
    is.na(max.size.ingest.um) ~ upper.tissue.trans.size.um,
    TRUE ~ pmin(max.size.ingest.um, upper.tissue.trans.size.um)
  )) %>% 
    # define environment-specific alpha parameters #
    mutate(alpha = case_when(environment == "Marine" ~ alpha.marine,
                             environment == "Freshwater" ~ alpha.freshwater),
           a.sa = case_when(environment == "Marine" ~ a.sa.marine,
                            environment == "Freshwater" ~ a.sa.freshwater),
           a.v = case_when(environment == "Marine" ~ a.v.marine,
                            environment == "Freshwater" ~ a.v.freshwater),
           a.m = case_when(environment == "Marine" ~ a.m.marine,
                           environment == "Freshwater" ~ a.m.freshwater),
           a.ssa = case_when(environment == "Marine" ~ a.ssa.marine,
                           environment == "Freshwater" ~ a.ssa.freshwater),
           ) %>% 
    # calculate effect threshold for particles
    mutate(EC_mono_p.particles.mL_trans = dose.particles.mL.master) %>% 
    mutate(mu.p.mono = 1) %>% #mu_x_mono is always 1 for particles to particles
    mutate(mu.p.poly_trans = mux_polyfnx_generalizable(a.x = alpha, #alpha for particles
                                         x_UL= x2M_trans, #upper ingestible size limit (width of particle)
                                         x_LL = x1M_set)) %>% 
    # polydisperse effect threshold for particles
    mutate(EC_poly_p.particles.mL_trans = (EC_mono_p.particles.mL_trans * mu.p.mono)/mu.p.poly_trans) %>% 
    #calculate CF_bio for all conversions
    mutate(CF_bio_trans = CFfnx(x1M = x1M_set,#lower size bin
                                x2M = x2M_trans, #upper translocatable
                                x1D = x1D_set, #default
                                x2D = x2D_set,  #default
                                a = alpha)) %>%  
    ## Calculate environmentally relevant effect threshold for particles
    mutate(EC_env_p.particles.mL_trans = EC_poly_p.particles.mL_trans * CF_bio_trans) %>%  #aligned particle effect concentraiton (1-5000 um)
    
    #### Surface area ERM ####
  ##--- environmental calculations ---###
  #calculate lower translocatable surface area
  mutate(x_LL_sa_trans = SAfnx(a = 0.5 * x1D_set, #length
                               b = 0.5 * x1D_set, #0.5 * R.ave * x1D_set, #width
                               c = 0.5 * x1D_set  #0.5 * R.ave * 0.67 * x1D_set #height
  )) %>%  
    #calculate upper translocatable surface area
    mutate(x_UL_sa_trans = SAfnx(a = 0.5 * x2M_trans, 
                                 b = 0.5 * x2M_trans, #width #0.5 * R.ave * x2M, 
                                 c = 0.5 * x2M_trans #heigth #0.5 * R.ave * 0.67 * x2M
    )) %>%  
    #calculate mu_x_poly (env) for surface area
    mutate(mu.sa.poly_trans = mux_polyfnx_generalizable(a.sa, x_UL_sa_trans, x_LL_sa_trans)) %>% 
    
    ##--- laboratory calculations ---###
    ## define mu_x_mono OR mu_x_poly (lab) for alignment to ERM  #
    #(note that if mixed particles were used, a different equation must be used)
    mutate(mu.sa.mono = case_when(
      polydispersity == "monodisperse" ~ particle.surface.area.um2, # use reported surface area in monodisperse
      polydispersity == "polydisperse" ~  mux_polyfnx_generalizable(a.x = a.sa, 
                                                      x_LL = particle.surface.area.um2.min,
                                                      x_UL = particle.surface.area.um2.max))) %>% 
    
    #calculate polydisperse effect concentration for surface area (particles/mL)
    mutate(EC_poly_sa.particles.mL_trans = (EC_mono_p.particles.mL_trans * mu.sa.mono)/mu.sa.poly_trans) %>%  
    #calculate environmentally realistic effect threshold
    mutate(EC_env_sa.particles.mL_trans = EC_poly_sa.particles.mL_trans * CF_bio_trans) %>% 
    
    ##### FOOD DILUTION ####
  # define upper size length for ingestion 
  mutate(x2M_ingest = case_when(is.na(max.size.ingest.um) ~ x2D_set, 
                                max.size.ingest.um < x2D_set ~ max.size.ingest.um,
                                max.size.ingest.um > x2D_set ~ x2D_set
  )) %>%  #set to 5,000 as upper limit or max size ingest, whichever is smaller
    # calculate effect threshold for particles
    mutate(EC_mono_p.particles.mL_ingest = dose.particles.mL.master) %>% 
    mutate(mu.p.mono = 1) %>% #mu_x_mono is always 1 for particles to particles
    mutate(mu.p.poly_ingest = mux_polyfnx_generalizable(a.x = alpha, #alpha for particles
                                          x_UL= x2M_ingest, #upper ingestible size limit
                                          x_LL = x1M_set)) %>% 
    # polydisperse effect threshold for particles
    mutate(EC_poly_p.particles.mL_ingest = (EC_mono_p.particles.mL_ingest * mu.p.mono)/mu.p.poly_ingest) %>% 
    #calculate CF_bio for all conversions
    mutate(CF_bio_ingest = CFfnx(x1M = x1M_set,#lower size bin
                                 x2M = x2M_ingest, #upper ingestible length
                                 x1D = x1D_set, #default
                                 x2D = x2D_set,  #default upper size range
                                 a = alpha)) %>%  
    ## Calculate environmentally relevant effect threshold for particles
    mutate(EC_env_p.particles.mL_ingest = EC_poly_p.particles.mL_ingest * CF_bio_ingest) %>%  #aligned particle effect concentraiton (1-5000 um)
    
    
    #### volume ERM ####
  ##--- environmental calculations ---###
  #calculate lower ingestible volume 
  mutate(x_LL_v_ingest = volumefnx_poly(length = x1D_set,
                                        width = x1D_set)) %>% 
    #calculate maximum ingestible volume 
    mutate(x_UL_v_ingest = volumefnx_poly(length = x2M_ingest, # length-limited
                                          #x2D_set, #upper definiton (accouunts for fibers) CONSERVATIVE
                                          width = x2M_ingest)) %>% #ingestion-limited
    # calculate mu.v.poly
    mutate(mu.v.poly_ingest = mux_polyfnx_generalizable(a.v, x_UL_v_ingest, x_LL_v_ingest)) %>% 
    ##--- laboratory calculations ---###
    ## define mu_x_mono OR mu_x_poly (lab) for alignment to ERM  #
    #(note that if mixed particles were used, a different equation must be used)
    mutate(mu.v.mono = case_when(
      polydispersity == "monodisperse" ~ particle.volume.um3, # use reported volume in monodisperse
      polydispersity == "polydisperse" ~ mux_polyfnx_generalizable(a.x = a.v, 
                                                     x_LL = particle.volume.um3.min,
                                                     x_UL = particle.volume.um3.max))) %>% 
    
    #calculate polydisperse effect concentration for volume (particles/mL)
    mutate(EC_poly_v.particles.mL_ingest = (EC_mono_p.particles.mL_ingest * mu.v.mono)/mu.v.poly_ingest) %>%  
    #calculate environmentally realistic effect threshold
    mutate(EC_env_v.particles.mL_ingest = EC_poly_v.particles.mL_ingest * CF_bio_ingest) %>% 
    
    ###### CLEANUP #####
  mutate(particles.mL.ox.stress = EC_env_sa.particles.mL_trans,
         particles.mL.food.dilution = EC_env_v.particles.mL_ingest) %>% 
    #### annotate studies in which particles are too big to be ingested or translocated for those ERMs (to be filtered in ssd functions)
    mutate(translocatable = ifelse(size.length.um.used.for.conversions > x2M_trans, 
                                   "not translocatable", 
                                   "translocatable")) %>% 
    mutate(ingestible = ifelse(size.length.um.used.for.conversions > x2M_ingest, 
                               "not ingestible", 
                               "ingestible")) %>% 
    
    #rowwise() %>%
    mutate(unique_id = row_number()) %>% 
    # mutate(unique_id = digest::digest(paste(across(everything()), collapse = "-"), algo = "md5")) %>%
    ungroup()
  
  return(aoc_MC_iter)
}

## derive deterministic aligned dataset
aoc_aligned_test <- alignmentFxn(df = tomex2.0_aoc_z_final)
```

### Model Wrapper
```{r}
#### Define Model Wrapper
model_wrapper <- function(params, iteration){
  
  # Print the current iteration
  print(paste("Iteration:", iteration))
  
  #unpack parameters
  # Ensure all extracted parameters are correctly coerced to numeric
  # Coerce to numeric directly while accessing the first element of potential list
  alpha.marine <- as.numeric(params$alpha.marine[1])
  a.sa.marine <- as.numeric(params$a.sa.marine[1])
  a.v.marine <- as.numeric(params$a.v.marine[1])
  a.m.marine <- as.numeric(params$a.m.marine[1])
  a.ssa.marine <- as.numeric(params$a.ssa.marine[1])
  alpha.freshwater <- as.numeric(params$alpha.freshwater[1])
  a.sa.freshwater <- as.numeric(params$a.sa.freshwater[1])
  a.v.freshwater <- as.numeric(params$a.v.freshwater[1])
  a.m.freshwater <- as.numeric(params$a.m.freshwater[1])
  a.ssa.freshwater <- as.numeric(params$a.ssa.freshwater[1])
  R_ave_water_marine <- as.numeric(params$R.ave.water.marine[1])
  R_ave_water_freshwater <- as.numeric(params$R.ave.water.freshwater[1])
  R_ave_sediment_marine <- as.numeric(params$R.ave.sediment.marine[1])
  R_ave_sediment_freshwater <- as.numeric(params$R.ave.sediment.freshwater[1])
  sim_beta_log10_body_length <- as.numeric(params$sim.beta.log10.body.length[1])
  sim_body_length_intercept <- as.numeric(params$sim.body.length.intercept[1])
  upper.tissue.trans.size.um <- as.numeric(params$upper.tissue.trans.size.um[1])
  
  
  ####### ---- RUN FUNCTIONS ---- ###
  ### Static with base parameters to ensure everything is working ###
  ## generate ToMEx 1.0 dataset ###
  aoc_setup <- ToMEx1.0fxn(R.ave.water.marine = R.ave.water.marine,
                           R.ave.water.freshwater = R.ave.water.freshwater,
                           R.ave.sediment.marine = R.ave.sediment.marine,
                           R.ave.sediment.freshwater = R.ave.sediment.freshwater,
                           beta_log10_body_length = beta_log10_body_length,
                           body_length_intercept = body_length_intercept) 
  
  ## generate ToMEx 2.0 dataset ###
  tomex2.0_aoc_z_final <- ToMEx2.0fxn(aoc_setup = aoc_setup,
                                      R.ave.water.marine = R.ave.water.marine,
                                      R.ave.water.freshwater = R.ave.water.freshwater,
                                      R.ave.sediment.marine = R.ave.sediment.marine,
                                      R.ave.sediment.freshwater = R.ave.sediment.freshwater,
                                      beta_log10_body_length = beta_log10_body_length,
                                      body_length_intercept = body_length_intercept)
  
  ### perform alignments
aoc_MC_iter <- alignmentFxn(df = tomex2.0_aoc_z_final)

  ##### Base Thresholds
  #filter out risk criteria (not done above)#
  aoc_risk_paper <- aoc_MC_iter %>%
    drop_na(effect.metric) %>%
    filter(tier_zero_tech_f == ("Red Criteria Passed"))
  
  # calculate thresholds for different environments
  marine_thresholds <- process_environment_data(aoc_risk_paper,
                                                "Marine", 
                                                upper.tissue.trans.size.um = upper.tissue.trans.size.um,
                                                x1D_set = x1D_set, 
                                                x2D_set = x2D_set)
  
  freshwater_thresholds <- process_environment_data(aoc_risk_paper, 
                                                    "Freshwater",
                                                    upper.tissue.trans.size.um = upper.tissue.trans.size.um,
                                                    x1D_set = x1D_set, 
                                                    x2D_set = x2D_set)
  
  # freshwater_marine_thresholds <- process_environment_data(aoc_risk_paper, 
  #                                                          c("Freshwater", "Marine"),
  #                                                          upper.tissue.trans.size.um = upper.tissue.trans.size.um,
  #                                                          x1D_set = x1D_set, 
  #                                                          x2D_set = x2D_set)
  
  
  ##### SAVE OUTPUT OF MONTE CARLO ######
  
  MC_results[[i]] <- list(particles_mL_ox_stress = aoc_MC_iter$particles.mL.ox.stress,
                          particles_mL_food_dilution = aoc_MC_iter$particles.mL.food.dilution,
                          unique_id = aoc_MC_iter$unique_id,
                          base_thresholds = list(
                            marine = marine_thresholds,
                            freshwater = freshwater_thresholds#,
                     #       freshwater_marine = freshwater_marine_thresholds
                          ))
} #close model_wrapper function
```

## Run Function
```{r}
# intialize dataframe to store MC results
MC_results <- vector("list", nrow(param_values))

## For loop to run MC simulation
for (i in 1:nrow(param_values)) {
  # Extract a single row of parameter samples as a list
  param_set <- param_values[i, ]
  
  # Run the model function
  MC_results[[i]] <- model_wrapper(param_set, i)
}

# save results
saveRDS(MC_results, "output/MC_results.rds")
```

## Load Monte Carlo Data
```{r}
#### START HERE IF WORKING FROM SAVED FILE ###
MC_results <- readRDS("output/MC_results.rds")
```

## Analysis
### Preparation
```{r}
# Extract and combine base_thresholds for each environment
all_thresholds_marine <- map(MC_results, ~ .x$base_thresholds$marine) %>% 
  bind_rows(.id = "simulation_id")

all_thresholds_freshwater <- map(MC_results, ~ .x$base_thresholds$freshwater) %>% 
  bind_rows(.id = "simulation_id")

# all_thresholds_freshwater_marine <- map(MC_results, ~ .x$base_thresholds$freshwater_marine) %>% 
#   bind_rows(.id = "simulation_id")

# Combine marine and freshwater data into one data frame with an additional "Environment" column
all_thresholds_combined <- bind_rows(
  mutate(all_thresholds_marine, Environment = "Marine"),
  mutate(all_thresholds_freshwater, Environment = "Freshwater"),
 # mutate(all_thresholds_freshwater_marine, Environment = "freshwater_marine"),
  .id = "simulation_id"
)

# Pivot data longer if the data structure requires it
# Assuming that your data might already be in a wide format and needs to be made long
all_thresholds_long <- all_thresholds_combined %>%
  pivot_longer(
    cols = -c(Tier, simulation_id, Environment),
    names_to = "Metric",
    values_to = "Value"
  )

```

### Summary
```{r}
# Calculate summary statistics
summary_stats_base_thresholds <- all_thresholds_long %>%
  group_by(Tier, Environment, Metric) %>%
  summarise(
    Mean = mean(Value, na.rm = TRUE),
    Median = median(Value, na.rm = TRUE),
    Std_Dev = sd(Value, na.rm = TRUE),
    Percentile_5th = quantile(Value, 0.05, na.rm = TRUE),
    Percentile_95th = quantile(Value, 0.95, na.rm = TRUE),
    N_sim = n(),
    .groups = 'drop'
  ) %>% 
  mutate_if(is.numeric, ~signif(., 3))

summary_stats_base_thresholds

```
```{r}
#save output
saveRDS(summary_stats_base_thresholds,
        "output/summary_stats_base_thresholds.rds")

# Output the summary statistics
print(summary_stats_base_thresholds)
```

### Plots
#### Function
```{r}
# define functino for making figures for marine and freshwater
MC_histograms_fxn <- function(long_data, environment,
                              #define custom limits for FD and tissue trans
                              food_limits = c(0.1, 1000), tissue_limits = c(100, 50000)){
  
  

  # long_data should be long data (see code for generating all_thresholds_long)
  
### Ggplot of SSD Tier Thresholds
# Prepare the data by calculating ERM and filtering
prepared_data <- long_data %>%  #all_thresholds_long %>%
  mutate(ERM = case_when(
    grepl("Tissue", Metric) ~ "Tissue Translocation",
    grepl("Food", Metric) ~ "Food Dilution"
  )) %>%
  filter(
    grepl("Default", Metric)
  ) %>% 
  filter(Environment == environment) %>% 
  drop_na(Value)


##### FOOD DILUTION (LEFT SIDE) ####
# Calculate median values for each group

median_data_food <- prepared_data %>%
  filter(ERM == "Food Dilution") %>% 
  group_by(Tier) %>%
  summarise(Median = median(Value, na.rm = T), .groups = 'drop')

# Assign specific colors based on Tier levels
tier_colors <- setNames(c("#b8e897", "#f0f062", "#f0a95f", "#f0514b"), levels(prepared_data$Tier))

  
MC_histograms_base_food <- ggplot(prepared_data %>% filter(ERM == "Food Dilution"),
                                  aes(x = Value)) +
  geom_density(aes(color = Tier), size = 1) +  # Map 'Tier' for color
  geom_histogram(aes(y = ..density.., fill = Tier), bins = 150, color = "black", alpha = 0.9, linewidth = 0.01) +  # Map 'Tier' for fill
  geom_vline(data = median_data_food, aes(xintercept = Median),
             linetype = "dotted", color = "black", size = 1.2) +
#  geom_text(data = median_data, aes(x = Median, y = 0, label = paste0("Median:", scientific(Median, 2))),
        #    hjust = +1.5, vjust = -1.5, color = "black", size = 6, fontface = "italic") +
  xlab("Particles/L") +
  scale_x_log10(#labels = scales::comma,
                labels = scales::scientific,
                limits = c(food_limits)) +
  facet_wrap(~ Tier, scales = "free_y", ncol = 1) +
  scale_fill_manual(values = tier_colors) +  # Apply manual colors for fill
  scale_color_manual(values = tier_colors)  # Apply manual colors for lines

library(ggdark)
MC_histograms_dark_food <- MC_histograms_base_food +
  dark_theme_bw(base_size = 20) +
  theme(
    strip.background = element_blank(),
    strip.placement = "outside",
    legend.position = "none",
    strip.text = element_blank(),  # This hides the facet wrap titles
    #  strip.text = element_text(face = "bold", size = 16, margin = margin(t = 1, b = 1)),
    panel.spacing = unit(0.1, "lines"),
    axis.text.y = element_blank(),
    axis.title.y = element_blank(),
    axis.ticks.y = element_blank()
  )

MC_histograms_light_food <- MC_histograms_base_food +
  theme_bw(base_size = 20) +
  theme(
    strip.background = element_blank(),
    strip.placement = "outside",
    legend.position = "none",
    strip.text = element_blank(),  # This hides the facet wrap titles
 #  strip.text = element_text(face = "bold", size = 16, margin = margin(t = 1, b = 1)),
    panel.spacing = unit(0.1, "lines"),
    axis.text.y = element_blank(),
    axis.title.y = element_blank(),
    axis.ticks.y = element_blank()
  )


##### Tissue Translocation (RIGHT SIDE) ####
# Calculate median values for each group

median_data_tissue <- prepared_data %>%
  filter(ERM == "Tissue Translocation") %>% 
  group_by(Tier) %>%
  summarise(Median = median(Value, na.rm = T), .groups = 'drop')


MC_histograms_base_tissue <- ggplot(prepared_data %>% filter(ERM == "Tissue Translocation"),
                                  aes(x = Value)) +
  geom_density(aes(color = Tier), size = 1) +  # Map 'Tier' for color
  geom_histogram(aes(y = ..density.., fill = Tier), bins = 150, color = "black", alpha = 0.9, linewidth = 0.01) +  # Map 'Tier' for fill
  geom_vline(data = median_data_tissue, aes(xintercept = Median),
             linetype = "dotted", color = "black", size = 1.2) +
#  geom_text(data = median_data, aes(x = Median, y = 0, label = paste0("Median:", scientific(Median, 2))),
        #    hjust = +1.5, vjust = -1.5, color = "black", size = 6, fontface = "italic") +
  xlab("Particles/L") +
  scale_x_log10(#labels = scales::comma,
                labels = scales::scientific,
                limits = tissue_limits) +
  facet_wrap(~ Tier, scales = "free_y", ncol = 1) +
  scale_fill_manual(values = tier_colors) +  # Apply manual colors for fill
  scale_color_manual(values = tier_colors)  # Apply manual colors for lines

library(ggdark)
MC_histograms_dark_tissue <- MC_histograms_base_tissue +
  dark_theme_bw(base_size = 20) +
  theme(
    strip.background = element_blank(),
    strip.placement = "outside",
    legend.position = "none",
    strip.text = element_blank(),  # This hides the facet wrap titles
    #  strip.text = element_text(face = "bold", size = 16, margin = margin(t = 1, b = 1)),
    panel.spacing = unit(0.1, "lines"),
    axis.text.y = element_blank(),
    axis.title.y = element_blank(),
    axis.ticks.y = element_blank()
  )

MC_histograms_light_tissue <- MC_histograms_base_tissue +
  theme_bw(base_size = 20) +
  theme(
    strip.background = element_blank(),
    strip.placement = "outside",
    legend.position = "none",
    strip.text = element_blank(),  # This hides the facet wrap titles
 #  strip.text = element_text(face = "bold", size = 16, margin = margin(t = 1, b = 1)),
    panel.spacing = unit(0.1, "lines"),
    axis.text.y = element_blank(),
    axis.title.y = element_blank(),
    axis.ticks.y = element_blank()
  )


#### COMBINE ####
MC_histograms_light <- ggarrange(MC_histograms_light_food, MC_histograms_light_tissue)
MC_histograms_dark <- ggarrange(MC_histograms_dark_food, MC_histograms_dark_tissue)

return(list(MC_histograms_light, MC_histograms_dark))
}
```

#### Generate Histograms
```{r}
### Generate plots
MC_histograms_marine <- MC_histograms_fxn(all_thresholds_long, "Marine",
                                          food_limits = c(0.00001, 500), tissue_limits = c(0.1, 100000))

MC_histograms_freshwater <- MC_histograms_fxn(all_thresholds_long, "Freshwater")

### Save plots
ggsave("output/MC_histograms_freshwater.png",
       MC_histograms_freshwater[[1]], 
      dpi = 300,
      width = 12, height = 9, units = "in")

ggsave("output/MC_histograms_marine.png",
       MC_histograms_marine[[1]], 
       dpi = 300,
       width = 12, height = 9, units = "in")


ggsave("output/presentation_Figs/MC_histograms.png",
       MC_histograms_marine[[2]], 
       dpi = 300,
       width = 10, height = 5,
       units = "in")

### Display plots
ggarrange(MC_histograms_marine[[1]], MC_histograms_freshwater[[1]],
          labels = c("Marine", "Freshwater"))
```

### Violin Plots
```{r}
### Ggplot of monte carlo SSD thresholds
all_thresholds_violin <- all_thresholds_long %>% 
  filter(grepl("Default", Metric)) %>%
  mutate(ERM = case_when(
    grepl("Tissue", Metric) ~ "Tissue Translocation",
    grepl("Food", Metric) ~ "Food Dilution")) %>% 
  ggplot(aes(x = Value, y = Tier, color = Environment, fill = Environment)) +
  #geom_boxplot(alpha = 0.5) +
  geom_violin(alpha = 0.9,
              draw_quantiles = c(0.05, 0.5, 0.95),
              color = "black") +
  cols4all::scale_fill_discrete_c4a_cat("carto.bold") +
  cols4all::scale_color_discrete_c4a_cat("carto.bold") +
  #geom_jitter() +
  scale_x_log10(name = "Particles/L (1 - 5000 um)",
                labels = scales::comma) +
  facet_grid(rows = vars(Environment), cols = vars(ERM),
             scales = "free") +
  theme_bw(base_size = 16) +
  theme(legend.position = "none")

ggsave("output/all_thresholds_violin.jpg",
       all_thresholds_violin, 
       dpi = 300,
       width = 10, height = 5,
       units = "in")

all_thresholds_violin
```




### Determine Coefficient of Variation
```{r}
##### Determine Coefficient of Variation for each output ###
### Step 1: Create a dataframe from the list
results_df <- do.call(rbind, lapply(MC_results, function(x) {
  data.frame(
    unique_id = x$unique_id,
    particles_L_ox_stress = x$particles_mL_ox_stress *1000,
    particles_L_food_dilution = x$particles_mL_food_dilution * 1000
  )
}))
skimr::skim(results_df)
```

```{r}
## Step 2:
# Calculate CoV
cov_results <- results_df %>%
  group_by(unique_id) %>%
  summarise(sd_particles_L_ox_stress = sd(particles_L_ox_stress, na.rm = TRUE),
            mean_particles_L_ox_stress = mean(particles_L_ox_stress, na.rm = TRUE),
            median_particles_L_ox_stress = median(particles_L_ox_stress, na.rm = TRUE),
            CoV_ox_stress = sd_particles_L_ox_stress / mean_particles_L_ox_stress,
            #food dilution
            sd_particles_L_food_dilution = sd(particles_L_food_dilution, na.rm = TRUE),
            mean_particles_L_food_dilution = mean(particles_L_food_dilution, na.rm = TRUE),
            median_particles_L_food_dilution = median(particles_L_food_dilution, na.rm = TRUE),
            CoV_food_dilution = sd_particles_L_food_dilution / mean_particles_L_food_dilution
  )

# View the results
skimr::skim(cov_results)
```

```{r}
# Step 3: merge with full dataset and deterministic dataset
aoc_MC <- merge(aoc_aligned_test, cov_results, by = "unique_id", all.x = TRUE)
aoc_MC <- left_join(aoc_MC, 
                    aoc_aligned_test %>% ungroup() %>%  
                      select(c(unique_id, particles.mL.food.dilution, particles.mL.ox.stress)) %>%
                      mutate(particles_L_food_dilution_deterministic = particles.mL.food.dilution * 1000,
                             particles_L_ox_stress_deterministic = particles.mL.ox.stress * 1000),
                    by = "unique_id"
) %>% 
  mutate(n_sim = n_sim)
#filter(size.length.um.used.for.conversions >= x1M_set) %>% #alignments not valid below 1 um, so filter them before they become problems later

#Save file for EcoTox Project
saveRDS(aoc_MC, "output/aoc_MC.rds")
```

### Quality Checks
```{r}
# summary stats to ensure MC was succesful 
aoc_MC %>% 
  select(c(particles_L_ox_stress_deterministic, median_particles_L_ox_stress,
           particles_L_food_dilution_deterministic, median_particles_L_food_dilution)) %>% 
  mutate(ox_stress_ratio = particles_L_ox_stress_deterministic / median_particles_L_ox_stress,
         food_dilution_ratio = particles_L_food_dilution_deterministic / median_particles_L_food_dilution)

#define limits
min_val <- min(aoc_MC$CoV_food_dilution, na.rm = TRUE)
max_val <- max(aoc_MC$CoV_food_dilution, na.rm = TRUE)

# You may want to add some padding around the min and max
pad <- (max_val - min_val) * 0.05
xlims <- c(min_val - pad, max_val + pad)

hist_cv_food <- aoc_MC %>% 
  ggplot(aes(x = CoV_food_dilution, fill = polydispersity)) +
  geom_histogram(position = "identity", alpha = 0.5) +
  xlim(xlims) +  # Set x limits
  scale_x_log10() +
  xlab("Coefficient of Variation (unitless)") +
  ylab("Toxicity Data Points")# +
#fill.type +
#  theme.light #+
#theme(axis.title.x = element_blank())

hist_cv_ox <- aoc_MC %>% 
  ggplot(aes(x = CoV_ox_stress, fill = polydispersity)) +
  geom_histogram(position = "identity", alpha = 0.5) +
  #fill.type +
  scale_x_log10() +
  # xlim(xlims) +  # Set x limits
  xlab("Coefficient of Variation (unitless)") +
  ylab("Toxicity Data Points") +
  # theme.light +
  theme(legend.position = "none",
        axis.title.y = element_blank())

CoV_plots <-  ggpubr::ggarrange(hist_cv_food, hist_cv_ox,
                                labels = c("A", "B"),
                                ncol = 2,
                                common.legend = TRUE, legend = "top")

ggsave(filename = "CoV_plots.jpg",
       dpi = 300,
       path = "../../output/Manuscript_Figs/", 
       plot = CoV_plots, width = 10, height = 8, units = "in")

CoV_plots
```

# Sensitivity Analysis (Sobol)
## Setup
### Define Wrapper
```{r}
model_wrapper_sobol <- function(params, N, save_interval = 500){
  
  # Initialize iteration counter
  iteration <- 1
  
  # Create list to store results
  sobol_results <- vector("list", N)
  
  # Report time
  start_time <- Sys.time()
  
  for (i in 1:N) {
    iter_start_time <- Sys.time()
    
    # Unpack parameters and ensure numeric conversion
    alpha.marine <- as.numeric(params$alpha.marine[1])
    a.sa.marine <- as.numeric(params$a.sa.marine[1])
    a.v.marine <- as.numeric(params$a.v.marine[1])
    a.m.marine <- as.numeric(params$a.m.marine[1])
    a.ssa.marine <- as.numeric(params$a.ssa.marine[1])
    alpha.freshwater <- as.numeric(params$alpha.freshwater[1])
    a.sa.freshwater <- as.numeric(params$a.sa.freshwater[1])
    a.v.freshwater <- as.numeric(params$a.v.freshwater[1])
    a.m.freshwater <- as.numeric(params$a.m.freshwater[1])
    a.ssa.freshwater <- as.numeric(params$a.ssa.freshwater[1])
    R_ave_water_marine <- as.numeric(params$R.ave.water.marine[1])
    R_ave_water_freshwater <- as.numeric(params$R.ave.water.freshwater[1])
    sim_beta_log10_body_length <- as.numeric(params$sim.beta.log10.body.length[1])
    sim_body_length_intercept <- as.numeric(params$sim.body.length.intercept[1])
    upper.tissue.trans.size.um <- as.numeric(params$upper.tissue.trans.size.um[1])
    
    ####### ---- RUN FUNCTIONS ---- ###
    # Generate ToMEx 1.0 dataset
    aoc_setup <- ToMEx1.0fxn(R.ave.water.marine = R_ave_water_marine,
                             R.ave.water.freshwater = R_ave_water_freshwater,
                             R.ave.sediment.marine = R.ave.sediment.marine,
                             R.ave.sediment.freshwater = R.ave.sediment.freshwater,
                             beta_log10_body_length = sim_beta_log10_body_length,
                             body_length_intercept = sim_body_length_intercept)
    
    # Generate ToMEx 2.0 dataset
    tomex2.0_aoc_z_final <- ToMEx2.0fxn(aoc_setup = aoc_setup,
                                        R.ave.water.marine = R_ave_water_marine,
                                        R.ave.water.freshwater = R_ave_water_freshwater,
                                        R.ave.sediment.marine = R.ave.sediment.marine,
                                        R.ave.sediment.freshwater = R.ave.sediment.freshwater,
                                        beta_log10_body_length = sim_beta_log10_body_length,
                                        body_length_intercept = sim_body_length_intercept)
    
    # Perform alignments
    aoc_MC_iter <- alignmentFxn(df = tomex2.0_aoc_z_final)
    
    # Filter out risk criteria and calculate thresholds for different environments
    aoc_risk_paper <- aoc_MC_iter %>%
      drop_na(effect.metric) %>%
      filter(tier_zero_tech_f == ("Red Criteria Passed"))
    
    marine_thresholds <- process_environment_data(aoc_risk_paper, "Marine",
                                                  upper.tissue.trans.size.um = upper.tissue.trans.size.um,
                                                  x1D_set = x1D_set, 
                                                  x2D_set = x2D_set)
    
    freshwater_thresholds <- process_environment_data(aoc_risk_paper, "Freshwater",
                                                      upper.tissue.trans.size.um = upper.tissue.trans.size.um,
                                                      x1D_set = x1D_set, 
                                                      x2D_set = x2D_set)
    
    # Store results in list
    sobol_results[[i]] <- list(
      base_thresholds = list(
        marine = marine_thresholds,
        freshwater = freshwater_thresholds
      )
    )
    
    # Save results every save_interval iterations
    if (i %% save_interval == 0) {
      filename <- paste0("output/sobol_results_iteration_", i, ".rds")
      saveRDS(sobol_results, file = filename)
      cat(sprintf("Saved results for iteration %d\n", i))
      flush.console()
    }
    
    iter_end_time <- Sys.time()
    iter_time <- iter_end_time - iter_start_time
    elapsed_time <- iter_end_time - start_time
    remaining_iterations <- N - i
    remaining_time <- (elapsed_time / i) * remaining_iterations
    
    cat(sprintf("Iteration %d/%d: Time per iteration = %.2f secs, Estimated remaining time = %.2f secs\n",
                i, N, iter_time, remaining_time))
  }
  
  total_time <- Sys.time() - start_time
  cat(sprintf("Total time for %d iterations: %.2f secs\n", N, total_time))
  
  return(sobol_results)
} #close model_wrapper function

```


### Define Matrix
```{r}
# Number of samples
N <- 300
matrices <- c("A", "B", "AB", "BA")
first <- total <- "azzini"

params <- c("alpha.marine", "a.sa.marine", "a.v.marine", "a.m.marine", "a.ssa.marine", 
            "alpha.freshwater", "a.sa.freshwater", "a.v.freshwater", "a.m.freshwater", "a.ssa.freshwater", 
            "R.ave.water.marine", "R.ave.water.freshwater",
          #  "R.ave.sediment.marine", "R.ave.sediment.freshwater",
            "sim.beta.log10.body.length", "sim.body.length.intercept",
            "upper.tissue.trans.size.um")

# Generate the Sobol' sequence
mat <- sobol_matrices(N = N, 
                      params = params,
                      type = "LHS"
                      )

# Convert to data.table
# Convert to data.table
mat <- data.table::as.data.table(mat)

# Transform each column to the specified probability distribution
mat[, alpha.marine := rnorm(.N,  mean = alpha_marine, sd = alpha.sd.marine)]
mat[, a.sa.marine := rtruncnorm(.N,  a = 0.0001, b = Inf, mean = a_sa.marine, sd = a.sa.sd.marine)]
mat[, a.v.marine := rtruncnorm(.N,  a = 0.0001, b = Inf, mean = a_v.marine, sd = a.v.sd.marine)]
mat[, a.m.marine := rtruncnorm(.N,  a = 0.0001, b = Inf, mean = a_m.marine, sd = a.m.sd.marine)]
mat[, a.ssa.marine := rtruncnorm(.N, a = 0.0001, b = Inf, mean = a_ssa.marine, sd = a.ssa.sd.marine)]
mat[, alpha.freshwater := rtruncnorm(.N,  a = 0.0001, b = Inf, mean = alpha_freshwater, sd = alpha.sd.freshwater)]
mat[, a.sa.freshwater := rtruncnorm(.N,  a = 0.0001, b = Inf, mean = a_sa.freshwater, sd = a.sa.sd.freshwater)]
mat[, a.v.freshwater := rtruncnorm(.N,  a = 0.0001, b = Inf, mean = a_v.freshwater, sd = a.v.sd.freshwater)]
mat[, a.m.freshwater := rtruncnorm(.N,  a = 0.0001, b = Inf, mean = a_m.freshwater, sd = a.m.sd.freshwater)]
mat[, a.ssa.freshwater := rtruncnorm(.N,  a = 0.0001, b = Inf, mean = a_ssa.freshwater, sd = a.ssa.sd.freshwater)]
mat[, R.ave.water.marine := rtruncnorm(.N, a = 0.0001, b= 0.9999, mean = R_ave.water.marine, sd = R.ave.water.marine.sd)]
mat[, R.ave.water.freshwater := rtruncnorm(.N, a = 0.0001, b= 0.9999, mean = R_ave.water.freshwater, sd = R.ave.water.freshwater.sd)]
mat[, sim.beta.log10.body.length := rnorm(.N, mean = beta_log10_body_length, sd = se_beta_log10_body_length)]
mat[, sim.body.length.intercept := rnorm(.N, mean = body_length_intercept, sd = se_body_length_intercept)]
#mat[, "R.ave.sediment.marine" := rtruncnorm(.N, a = 0.0001, b= 0.9999, mean = 0.75, sd = 0.30)]
#mat[, "R.ave.sediment.freshwater" := rtruncnorm(.N, a = 0.0001, b= 0.9999, mean = 0.70, sd = 0.33)]



# Parameters from logistic regression model
beta_0 <- 1.308344
beta_1 <- -0.01468148
se_beta_0 <- 0.3963612
se_beta_1 <- 0.006657993

# Simulate beta_0 and beta_1
sim_beta_0 <- rnorm(nrow(mat) * 1.2, mean = beta_0, sd = se_beta_0)
sim_beta_1 <- rnorm(nrow(mat) * 1.2, mean = beta_1, sd = se_beta_1)

# Calculate X50 for each simulation
sim_X50 <- -sim_beta_0 / sim_beta_1

# Truncate distribution to not fall below 0
upper.tissue.trans.size.um_samples <- sim_X50 %>% 
  data.frame() %>% 
  filter(. > x1M_set) %>% 
  slice(1:nrow(mat))

upper.tissue.trans.size.um_samples <- as.numeric(upper.tissue.trans.size.um_samples$.)

mat[, upper.tissue.trans.size.um := upper.tissue.trans.size.um_samples]

# Convert the data.table to a data.frame
mat <- as.data.frame(mat)

# save mat (necessary for stats later)
saveRDS(mat, file = "output/mat.rds")

skimr::skim(mat)
```

### Run Sobol
```{r message=FALSE}
# 2. Run the Model for Each Sample Set: Run your model for each set of parameter samples.
# Initialize a list to store the results
# Initialize the results list and counter
sobol_results <- vector("list", nrow(mat))
N <- as.integer(nrow(mat))
N

# Loop through each row of the parameter matrix
for (i in 1:N) {
  param_set <- mat[i, ]
  
    # Perform the model wrapping with error handling
  tryCatch({
    sobol_results[i] <- model_wrapper_sobol(param_set, N = 1, save_interval = 2) # N=1 because each call handles one set of params
  }, error = function(e) {
    cat(sprintf("Error at iteration %d: %s\n", i, e$message))
    flush.console()
  })
  
  # Print progress and flush the console output
  cat(sprintf("Processing iteration %d\n", i))
  flush.console()
}

# Save the final results
saveRDS(sobol_results, file = "output/sobol_results.rds")
```

### Read in previously run Sobol
```{r}
#### start here from saved file ###
sobol_results <- readRDS("output/sobol_results.rds")
```


## Analysis
### Histograms
```{r}
# Extract and combine marine and freshwater thresholds
all_thresholds_marine_sobol <- map_dfr(sobol_results, ~ .x$base_thresholds$marine, .id = "simulation_id")
all_thresholds_freshwater_sobol <- map_dfr(sobol_results, ~ .x$base_thresholds$freshwater, .id = "simulation_id")

# Combine marine and freshwater data into one data frame with an additional "Environment" column
all_thresholds_sobol_combined <- bind_rows(
  mutate(all_thresholds_marine_sobol, Environment = "Marine"),
  mutate(all_thresholds_freshwater_sobol, Environment = "Freshwater"),  .id = "simulation_id")

# Pivot data longer if the data structure requires it
# Assuming that your data might already be in a wide format and needs to be made long
all_thresholds_sobol_long <- all_thresholds_sobol_combined %>%
  pivot_longer(
    cols = -c(Tier, simulation_id, Environment),
    names_to = "Metric",
    values_to = "Value"
  )


# 3. Extract the Output of Interest: Extract the relevant output from each model run (marine and freshwater thresholds).
# Convert results to a numeric vector
Y_marine_food_T3 <- as.numeric(all_thresholds_marine_sobol %>% filter(Tier == "Tier3") %>%  pull(`Food Dilution (Default)`))
Y_freshwater_food_T3 <- as.numeric(all_thresholds_freshwater_sobol %>% filter(Tier == "Tier3") %>%  pull(`Food Dilution (Default)`))

Y_marine_tissue_T3 <- as.numeric(all_thresholds_marine_sobol %>% filter(Tier == "Tier3") %>%  pull(`Tissue Translocation (Default)`))
Y_freshwater_tissue_T3 <- as.numeric(all_thresholds_freshwater_sobol %>% filter(Tier == "Tier3") %>%  pull(`Tissue Translocation (Default)`))


#histogram of results
hist(Y_marine_food_T3)
hist(Y_freshwater_food_T3)
```
#### GGplot Histograms
```{r}
Freshwater_histograms_sobol <- MC_histograms_fxn(all_thresholds_sobol_long, "Freshwater",
                                                 food_limits = c(0.01, 5000), tissue_limits = c(100, 50000))

Marine_histograms_sobol <- MC_histograms_fxn(all_thresholds_sobol_long, "Marine",
                                             food_limits = c(0.000001, 20), tissue_limits = c(0.5, 5000))

### Save plots
ggsave("output/MC_histograms_freshwater_sobol.png",
      Freshwater_histograms_sobol[[1]], 
      dpi = 300,
      width = 12, height = 8, units = "in")

ggsave("output/MC_histograms_marine_sobol.png",
       Marine_histograms_sobol[[1]], 
       dpi = 300,
       width = 12, height = 8, units = "in")


ggsave("output/presentation_Figs/MC_histograms_sobol.png",
       Marine_histograms_sobol[[2]], 
       dpi = 300,
       width = 12, height = 4,
       units = "in")

### Display plots
ggarrange(Marine_histograms_sobol[[1]], Freshwater_histograms_sobol[[1]],
          labels = c("Marine", "Freshwater"))
```

#### Summary Statistics
```{r}
summary_stats_base_thresholds_sobol <- as.tibble(all_thresholds_sobol_long) %>% 
  filter(grepl("5th|95th|Default", Metric)) %>% 
  group_by(Tier, Environment, Metric) %>%
  summarise(
    Mean = mean(Value, na.rm = TRUE),
    Median = median(Value, na.rm = TRUE),
    Std_Dev = sd(Value, na.rm = TRUE),
    Percentile_5th = quantile(Value, 0.05, na.rm = TRUE),
    Percentile_95th = quantile(Value, 0.95, na.rm = TRUE),
    N_sim = n(),
    .groups = 'drop'
  ) %>% 
  mutate_if(is.numeric, ~signif(., 3)) %>% 
  drop_na()

write.csv(summary_stats_base_thresholds_sobol, "output/summary_stats_base_thresholds_sobol.csv")

summary_stats_base_thresholds_sobol %>% mutate(across(where(is.numeric), ~scales::comma(.)))
```
##### Simple Summary
```{r}
simple_table <- summary_stats_base_thresholds_sobol %>% 
   mutate(across(where(is.numeric), ~scales::comma(.))) %>% 
  mutate(value = paste0(Mean, " (", Percentile_5th, " to ", Percentile_95th, ")")) %>% 
  filter(grepl("Default", Metric)) %>% 
  select(Environment, Metric, value, Tier) %>% 
  pivot_wider(values_from = value,
              names_from = Tier) %>% 
  as.data.frame() %>% 
  t() %>% 
   as.data.frame() %>%  # Convert the transposed matrix back to a data frame
  rownames_to_column(var = "Variable")  # Convert row names to a column for clarity

# Set the first row as the column names
colnames(simple_table) <- simple_table[1,]
simple_table <- simple_table[-1,]
rownames(simple_table) <- NULL

write.csv(simple_table, "output/simple_table.csv")

simple_table
```
##### Percent RSD
```{r}
simple_RSD <- summary_stats_base_thresholds_sobol %>% 
  #mutate(across(where(is.numeric), ~scales::comma(.))) %>% 
  mutate(percentRSD = (100* Std_Dev) /  Mean) %>% 
  filter(grepl("Default", Metric)) %>% 
  select(Environment, Metric, percentRSD, Tier) %>% 
  pivot_wider(values_from = percentRSD,
              names_from = Tier) %>% 
  as.data.frame() %>% 
  t() %>% 
   as.data.frame() %>%  # Convert the transposed matrix back to a data frame
  rownames_to_column(var = "Variable")  # Convert row names to a column for clarity

# Set the first row as the column names
colnames(simple_RSD) <- simple_RSD[1,]
simple_RSD <- simple_RSD[-1,]
rownames(simple_RSD) <- NULL

write.csv(simple_RSD, "output/simple_RSD.csv")

simple_RSD
```


### Scatterplots
```{r}
#stopped MC at iteration = 2037, so need to subset mat to that value
#mat2 <- mat[1:2037,]
mat2 <- mat

#scatter plot of parameter interactions
scatter_marine_food_t3 <- sensobol::plot_scatter(data = mat2, N = N, Y = Y_marine_food_T3, params = params, method = "bin") + ggtitle("Scatter Plot: Marine Food")

scatter_freshwater_food_t3 <- sensobol::plot_scatter(data = mat2, N = N, Y = Y_freshwater_food_T3,
                                                     params = params, method = "bin") + ggtitle("Scatter Plot: Freshwater Food")

scatter_marine_tissue_t3 <- sensobol::plot_scatter(data = mat2, N = N, Y = Y_marine_tissue_T3, params = params, method = "bin") + ggtitle("Scatter Plot: Marine Tissue")

scatter_freshwater_tissue_t3 <- sensobol::plot_scatter(data = mat2, N = N, Y = Y_freshwater_tissue_T3, params = params, method = "bin") + ggtitle("Scatter Plot: Freshwater Tissue")
 
# Arrange the plots into a 2x2 matrix with a common legend
scatterplots <- combined_plot <- ggpubr::ggarrange(
  scatter_marine_food_t3, scatter_freshwater_food_t3,
  scatter_marine_tissue_t3, scatter_freshwater_tissue_t3,
  ncol = 2, nrow = 2, common.legend = TRUE, legend = "bottom"
)

ggsave(filename = "sobol_scatterplots.jpg",
       dpi = 300,
       path = "output", 
       plot = scatterplots, width = 14, height = 10, units = "in")

scatterplots
```

### Multi-Scatter
#### Food Dilution
```{r}
multi_food <- plot_multiscatter(data = mat2, 
                  N = N,
                  Y = Y_marine_food_T3, 
                  params = c("alpha.marine", "a.v.marine", "R.ave.water.marine", "sim.body.length.intercept"))

multi_food
```

#### Tissue Translocation
```{r}
multi_tissue <- plot_multiscatter(data = mat2, 
                                N = N,
                                Y = Y_freshwater_tissue_T3, 
                                params = c("R.ave.water.freshwater", "a.sa.freshwater", "upper.tissue.trans.size.um"))

multi_tissue
```

### Quality Checks
```{r}
#### checks ##
# Check for NAs in parameter values
any(is.na(mat))

# Check for NAs in model output
any(is.na(Y_marine_food_T3))
any(is.na(Y_freshwater_food_T3))
any(is.na(Y_marine_tissue_T3))
any(is.na(Y_freshwater_tissue_T3))

# Check variance of parameters
apply(mat, 2, var)

# Check variance of outputs
var(Y_marine_food_T3)
var(Y_freshwater_tissue_T3)
```

```{r}
# Check for NAs in model output
valid_indices <- complete.cases(Y_marine_food_T3, Y_marine_tissue_T3, Y_freshwater_tissue_T3, Y_freshwater_food_T3)
Y_marine_food_T3_valid <- Y_marine_food_T3[valid_indices]
Y_marine_tissue_T3_valid <- Y_marine_tissue_T3[valid_indices]
Y_freshwater_food_T3_valid <- Y_freshwater_food_T3[valid_indices]
Y_freshwater_tissue_T3_valid <- Y_freshwater_tissue_T3[valid_indices]
mat_valid <- mat[valid_indices, ]
```

```{r}
# Update N to the number of valid samples
N_valid <- as.numeric(nrow(mat_valid))
#N must be integer, with Y = 15 * N
N <- N_valid / ncol(mat)
N <- as.integer(N)
N
length_y_rows <- N * ncol(mat)
length_y_rows
#4860
```

### Calculate Sobol Indices
```{r}
R <- 4000 #bootstrap iterations

ind.marine_food_T3 <- sobol_indices(Y = Y_marine_food_T3_valid[1:length_y_rows], 
                                    N = ncol(mat), params = params, boot = TRUE, R = R)

ind.marine_tissue_T3 <- sobol_indices(Y = Y_marine_tissue_T3_valid[1:length_y_rows],
                                      N = ncol(mat), params = params, boot = TRUE, R = R)

ind.freshwater_food_T3 <- sobol_indices(Y = Y_freshwater_food_T3_valid[1:length_y_rows],
                                        N = ncol(mat), params = params, boot = TRUE, R = R)

ind.freshwater_tissue_T3 <- sobol_indices(Y = Y_freshwater_tissue_T3_valid[1:length_y_rows],
                                          N = ncol(mat), params = params, boot = TRUE, R = R)
```

#### Display Indices
```{r}
#The output of sobol_indices() is an S3 object of class sensobol, with the results stored in the component results. To improve the visualization of the object, we set the number of digits in each numerical column to 3:
cols <- colnames(ind.marine_food_T3$results)[1:5]
ind.marine_food_T3$results[,(cols):=round(.SD,3),.SDcols=(cols)]

ind.marine_food_T3
```

```{r}
#WecanalsocomputetheSobol indicesofadummyparameter, i.e., aparameter thathas no influence on themodel output, to estimate thenumerical approximationerror. This will beused later onto identifyparameterswhose contributionto theoutputvariance is less than the approximationerror, and therefore cannot be considered influential. Like sobol_indices(),thefunctionsobol_dummy()allowstoobtainpointestimates(thedefault) orbootstrapestimates. Inthisexampleweusethelatteroption:
ind.marine_food_T3_dummy <- sobol_dummy(Y = Y_marine_food_T3_valid[1:length_y_rows], N = N, params = params, boot = TRUE, R = R)

ind.marine_tissue_T3_dummy <- sobol_dummy(Y = Y_marine_tissue_T3_valid[1:length_y_rows], N = N, params = params, boot = TRUE, R = R)
ind.freshwater_food_T3_dummy <- sobol_dummy(Y = Y_freshwater_food_T3_valid[1:length_y_rows], N = N, params = params, boot = TRUE, R = R)
ind.freshwater_tissue_T3_dummy <- sobol_dummy(Y = Y_freshwater_tissue_T3_valid[1:length_y_rows], N = N, params = params, boot = TRUE, R = R)


plot(ind.marine_food_T3, dummy = ind.marine_food_T3_dummy)
plot(ind.marine_tissue_T3, dummy = ind.marine_tissue_T3_dummy)
```

```{r}
# 4. Calculate Sobol' Indices: Calculate the Sobol' indices using the results.
# Function to plot Sobol' indices
plot_indices <- function(indices, title) {
  indices$results %>% 
    mutate(sensitivity = case_when(sensitivity == "Si" ~ "First-Order Sobol Index",
                                   sensitivity == "Ti" ~ "Total-Order Sobol Index")) %>% 
  ggplot(aes(y = parameters, x = original, fill = sensitivity)) +
    geom_bar(stat = "identity", position = "dodge") +
    geom_errorbarh(aes(xmin = low.ci, xmax = high.ci), 
                   position = position_dodge(0.9), height = 0.25,
                   color = "black") +
    labs(#title = title,
         #y = "Parameters",
         x = "Sobol' Index") +
    scale_y_discrete(position = "right") +  # Duplicate y-axis on the right side
    theme_minimal(base_size = 14) +
    theme(legend.title = element_blank())
}
```

```{r}
# Plot Sobol' indices for marine thresholds
sobol_plot_marine_food <-  plot_indices(ind.marine_food_T3, "Sobol' Indices for Marine Food Thresholds")

# Plot Sobol' indices for freshwater thresholds
sobol_plot_freshwater_tissue <-  plot_indices(ind.freshwater_tissue_T3, "Sobol' Indices for Freshwater Tissue Thresholds")

# Plot Sobol' indices for freshwater thresholds
sobol_plot_marine_tissue <-  plot_indices(ind.marine_tissue_T3, "Sobol' Indices for Marine Tissue Thresholds")

# Plot Sobol' indices for freshwater thresholds
sobol_plot_freshwater_food <-  plot_indices(ind.freshwater_food_T3, "Sobol' Indices for Freshwater Food Thresholds")

### Arrange sobol plots
indices <- ggarrange(sobol_plot_marine_food + labs(title = "Food Dilution (Marine)") +
            theme(axis.title.x = element_blank(), axis.title.y = element_blank(), plot.title = element_text(hjust = 0.5)),
          sobol_plot_freshwater_food + labs(title = "Food Dilution (Freshwater)") +
            theme(axis.text.y = element_blank(), axis.title.y = element_blank(),axis.title.x = element_blank(), plot.title = element_text(hjust = 0.5)),
          sobol_plot_marine_tissue + labs(title = "Tissue Translocation (Marine)") + theme(axis.title.y = element_blank(), plot.title = element_text(hjust = 0.5)),# + labs(y = "Tissue Translocation"),
          sobol_plot_freshwater_tissue + labs(title = "Tissue Translocation (Freshwater)") + theme(axis.text.y = element_blank(), axis.title.y = element_blank(), plot.title = element_text(hjust = 0.5)),
          common.legend = TRUE, legend = "bottom",
          labels = c("A", "B", "C", "D"))

ggsave(filename = "sobol_indices.jpg",
       dpi = 300,
       path = "output/", 
       plot = indices, width = 12, height = 9, units = "in")


indices
```

