---
title: "PSSD_WiggerData_updated"
output:
  html_document:
    code_folding: hide
    theme: journal
    toc: yes
    toc_float: yes
    toc_depth: 6
    number_sections: true
  word_document:
    toc: yes
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(fig.width=10, fig.height=8, fig.path='output/figures',
                      warning=FALSE, message=FALSE,time_it = TRUE) 
```

#Setup
##Libraries
```{r}
library(xlsx)
library(psych)
library(mc2d)
library(trapezoid)
library(tidyverse)
library(reshape2)
library(ggdark)
```
--
This script uses ToMEx 2.0 data.


 Code based on PSSD and PNEC calculations using the tool PSSD+ developed by Wigger et al. (2019)
 -------------------------------------------------------------------------------------------------

 Associated publication: Systematic consideration of parameter uncertainty and variability in
                         probabilistic species sensitivity distributions
 Authors: Henning Wigger, Delphine Kawecki, Bernd Nowack and V?ronique Adam

 Institute: Empa, Swiss Federal Laboratories for Materials Science and Technology,
            Technology and Society Laboratory, Lerchenfeldstrasse 5, 9014 St. Gallen, Switzerland

 submitted to Integrated Environmental Assessment and Management in December 2018

 -------------------------------------------------------------------------------------------------

This script was updated by Dr. Win Cowger to run without an error that amplified simulation magnitude based on a value (Win - fill in details).

## Parameters and Functions
```{r}
#### Main parameters ####

# number of simulations for the triangular distributions of the endpoints and assessment factors
SIM <- 10000

# coefficient of variation for the endpoint distributions
CV.DP <- 0.3

# coefficient of variation for the assessment factor distributions
CV.UF <- 0.5

## read functions
source("PSSD/do.pssd.r")
source("PSSD/rmore.r")
```

## Data Preparation
```{r}
#prepped in ToMEx2_EcoTox.Rmd
tier1_2_large <- readRDS("../output/tier1_2_large.rds")

t2 <- tier1_2_large %>% 
  distinct(dose_new, .keep_all = TRUE)#  %>% 
  #filter(Species %in% c("Daphnia magna", "Oryzias latipes"))

# prep dataframe for tox datapoints
DP_1_large_volume <-  t2 %>% 
  #mutate(tier = "A") %>% 
  mutate(id = row_number()) %>% 
  mutate(particles_L = EC_env_v.particles.mL_ingest * 1000) %>% 
  select(id, Species, particles_L) %>% 
  acast(id ~ Species, value.var = "particles_L") %>% 
  Matrix(sparse = TRUE) # makes a smaller object than using as.matrix()

#prep uncertainty factors for conversion from acute to chronic values
UFt_1_large_volume <- t2 %>% 
  mutate(id = row_number()) %>% 
  select(id, Species, af.time) %>% 
  acast(id ~ Species, value.var = "af.time")

# uncertainty factors for conversion of dose descriptors to NOEC
UFdd_1_large_volume <- t2 %>% 
  mutate(id = row_number()) %>% 
  select(id, Species, af.noec) %>% 
  acast(id ~ Species, value.var = "af.noec")

# dose descriptor
doseDescriptor_1_large_volume <- t2 %>% 
  mutate(id = row_number()) %>% 
  select(id, Species, effect.metric) %>% 
  acast(id ~ Species, value.var = "effect.metric")

# Polymer data as matrix
polymer_1_large_volume <- t2 %>% 
  mutate(id = row_number()) %>% 
  select(id, Species, poly_f) %>% 
  acast(id ~ Species, value.var = "poly_f")

#Shape data
shape_1_large_volume <- t2 %>% 
  mutate(id = row_number()) %>% 
  select(id, Species, shape_f) %>% 
  acast(id ~ Species, value.var = "shape_f")

#environment
env_1_large_volume <- t2 %>% 
  mutate(id = row_number()) %>% 
  select(id, Species, environment) %>% 
  acast(id ~ Species, value.var = "environment")

# group
group_1_large_volume <- t2 %>% 
  mutate(id = row_number()) %>% 
  select(id, Species, Group) %>% 
  acast(id ~ Species, value.var = "Group")

# read toxicity data
# data points from the literature
DP <- DP_1_large_volume

test <- as.matrix(DP)

# uncertainty factors for conversion from acute to chronic values
UFt  <- UFt_1_large_volume
# uncertainty factors for conversion of dose descriptors to NOEC
UFdd <- UFdd_1_large_volume
```

```{r include=FALSE}
#import Wigger dataset#
#### read input data ####
library(readxl)

# requires species values as endpoint value 
Datapoints <- DP

# requires assessment factors per species endpoint value - AF for time
UFt <- UFt

# AF for endpoint
UFdd <- UFdd

# Polymers used for each endpoint
Polymer <- polymer_1_large_volume

# Shape of microplastics for each endpoint
Shape <- shape_1_large_volume

# Dose descriptors of toxicity
DDescriptor <- doseDescriptor_1_large_volume

########################

# run 10 separate pSSDs if necessary, why, this is probably a ram issue?
# Initialize an empty list to store the results
pSSD_list <- vector("list", 10)

# Loop through each iteration
for(i in 1:10) {
  pSSD_list[[i]] <- do.pSSD(DP = Datapoints,
                            UFt = UFt,
                            UFdd = UFdd,
                            SIM/10, 
                            CV.DP,
                            CV.UF)
}

# Combine all the results into a single object
pSSD <- do.call(cbind, pSSD_list)

save(pSSD, file = "PSSD/updated/pSSD_tier1_2_large.RData")

corr.endpoints <- Datapoints/(UFdd*UFt)
```

#Plotting
## Original code from Wigger et al
````{r}
library(psych) # To use the geometric.mean function

#open a pdf file to start plotting
{ pdf(file = "PSSD_tier1_2_large.pdf",
      height = 4,
      width  = 5,
      pointsize = 10)
  #some parameters for margins
  par(mfrow = c(1,1), mar = c(4,4.5,3,1), mgp = c(2.5,1,0), xpd = F)
  
  # empty plot with annotations
  plot(c(-4, 8), c(0,1),
       main="PSSD of microplastics in aquatic environments",
       xlab = "NOEC (particles/L)", ylab = "Cumulative probability",
       type = "n", axes = F)
  axis(1, at = c( -4, -2, 0, 2, 4, 6, 8), 
       labels = expression(10^-4, 10^-2, 1, 10^2, 10^4, 10^6, 10^8), lwd = 0.5)
  axis(2, lwd = 0.5)
  abline(h = seq(0, 1, 0.2), lty = 1, col = "gray90", lwd = 0.5)
  abline(v = seq(-4, 8, 1), lty = 1, col = "gray90", lwd = 0.5)
  box(lwd = 0.5)
  
  # plot some ecdfs
  iv <- seq(-3, 7, 0.001)
  ECDF.data <- matrix(NA, 10000, length(iv))
  
  for(i in 1:10000){
    # calculate the ecdf function
    the.ecdf.f <- ecdf(log(pSSD[,i], base = 10))
    ECDF.data[i,] <- the.ecdf.f(iv)
  }
  
  # to plot quantiles of ECDFs:
  polygon(c(iv,rev(iv)),
          c(apply(ECDF.data,2,function(x){ min(x) }),
            rev(apply(ECDF.data,2,function(x){ max(x) }))),
          border = NA, col = adjustcolor("coral", alpha.f = 0.2))
  
  polygon(c(iv,rev(iv)),
          c(apply(ECDF.data,2,function(x){ quantile(x,0.05) }),
            rev(apply(ECDF.data,2,function(x){ quantile(x,0.95) }))),
          border = NA, col = adjustcolor("coral", alpha.f = 0.5))
  
  polygon(c(iv,rev(iv)),
          c(apply(ECDF.data,2,function(x){ quantile(x,0.25) }),
            rev(apply(ECDF.data,2,function(x){ quantile(x,0.75) }))),
          border = NA, col = adjustcolor("coral"))
  
  
  # calculate and plot the median for each x value of every ecdf pSSD
  # lines(iv, apply(ECDF.data, 2, median), col = "firebrick4", lwd = 2)
  # 
  # # OR calculate and plot the mean for each x value of every ecdf pSSD
  lines(iv, apply(ECDF.data, 2, mean), col = "firebrick4", lwd = 2)
  # 
  # then calculate the geometric mean of the endpoints to show on curve
  
  # deterministic values of NOEC (for plotting)
  NOEC.det <- Datapoints /( UFd * UFt)
  
  # calculate the geometric mean
  NOEC.gmean <- apply(NOEC.det, 2, function(x) geometric.mean(x, na.rm =T ))
  
  # calculate where to put the endpoints on the y axis
  #prop <- 1:ncol(Datapoints)/(ncol(Datapoints)+1)
  prop <- rep(NA, length(NOEC.gmean))
  for (i in 1:length(NOEC.gmean)) {
    prop[i] <- i/(length(NOEC.gmean))-(1/(length(NOEC.gmean))/2)
  }
  
  ind <- rank(NOEC.gmean)
  ind[4] <-4
  ind[12]<-21
  ind[14]<-23
  
  # overlap for two species
  library(stringr)
  
  # whose with additives  
  for (sp in 1:ncol(Datapoints)){
    for (i in 1:nrow(Datapoints)){
      if (str_detect((Shape[i,sp]), "fiber") && !is.na(Shape[i,sp])){
        if (str_detect(Polymer[i,sp], "PET") == T){
          points(x = log10(NOEC.det[i,sp]),
                 y = prop[ind[sp]],
                 pch = 4, cex = 1, col = "green4")
        } else if (str_detect(Polymer[i,sp], "PP") == T){
          points(x = log10(NOEC.det[i,sp]),
                 y = prop[ind[sp]],
                 pch = 4, cex = 1, col = "blue")
        }
      }
      else if (str_detect((Shape[i,sp]), "sphere") && !is.na(Shape[i,sp])){
        if (str_detect(Polymer[i,sp], "PS") == T){
          points(x = log10(NOEC.det[i,sp]),
                 y = prop[ind[sp]],
                 pch = 16, cex = 1, col = "gray18")
        } else if (str_detect(Polymer[i,sp], "PET") == T){
          points(x = log10(NOEC.det[i,sp]),
                 y = prop[ind[sp]],
                 pch = 16, cex = 1, col = "green4") 
        } else if (str_detect(Polymer[i,sp], "PVC") == T){
          points(x = log10(NOEC.det[i,sp]),
                 y = prop[ind[sp]],
                 pch = 16, cex = 1, col = "skyblue")
        } else if (str_detect(Polymer[i,sp], "PE") == T){
          points(x = log10(NOEC.det[i,sp]),
                 y = prop[ind[sp]],
                 pch = 16, cex = 1, col = "darkgray")
        } else if (str_detect(Polymer[i,sp], "PLA") == T){
          points(x = log10(NOEC.det[i,sp]),
                 y = prop[ind[sp]],
                 pch = 16, cex = 1, col = "lightgreen")  
        } else if (str_detect(Polymer[i,sp], "PA") == T){
          points(x = log10(NOEC.det[i,sp]),
                 y = prop[ind[sp]],
                 pch = 16, cex = 1, col = "red")
        } else if (str_detect(Polymer[i,sp], "PUR") == T){
          points(x = log10(NOEC.det[i,sp]),
                 y = prop[ind[sp]],
                 pch = 16, cex = 1, col = "yellow")
        } else if (str_detect(Polymer[i,sp], "PP") == T){
          points(x = log10(NOEC.det[i,sp]),
                 y = prop[ind[sp]],
                 pch = 16, cex = 1, col = "blue")  
        } else if (str_detect(Polymer[i,sp], "AF") == T){
          points(x = log10(NOEC.det[i,sp]),
                 y = prop[ind[sp]],
                 pch = 16, cex = 1, col = "mediumpurple")
        }
      }
      else if (str_detect((Shape[i,sp]), "fragment") && !is.na(Shape[i,sp])){
        if (str_detect(Polymer[i,sp], "PS") == T){
          points(x = log10(NOEC.det[i,sp]),
                 y = prop[ind[sp]],
                 pch = 15, cex = 1, col = "gray18")
        } else if (str_detect(Polymer[i,sp], "PET") == T){
          points(x = log10(NOEC.det[i,sp]),
                 y = prop[ind[sp]],
                 pch = 15, cex = 1, col = "green4") 
        } else if (str_detect(Polymer[i,sp], "PVC") == T){
          points(x = log10(NOEC.det[i,sp]),
                 y = prop[ind[sp]],
                 pch = 15, cex = 1, col = "skyblue")
        } else if (str_detect(Polymer[i,sp], "PE") == T){
          points(x = log10(NOEC.det[i,sp]),
                 y = prop[ind[sp]],
                 pch = 15, cex = 1, col = "darkgray")
        } else if (str_detect(Polymer[i,sp], "PLA") == T){
          points(x = log10(NOEC.det[i,sp]),
                 y = prop[ind[sp]],
                 pch = 15, cex = 1, col = "lightgreen")  
        } else if (str_detect(Polymer[i,sp], "PA") == T){
          points(x = log10(NOEC.det[i,sp]),
                 y = prop[ind[sp]],
                 pch = 15, cex = 1, col = "red")
        } else if (str_detect(Polymer[i,sp], "PUR") == T){
          points(x = log10(NOEC.det[i,sp]),
                 y = prop[ind[sp]],
                 pch = 15, cex = 1, col = "yellow")  
        } else if (str_detect(Polymer[i,sp], "PP") == T){
          points(x = log10(NOEC.det[i,sp]),
                 y = prop[ind[sp]],
                 pch = 15, cex = 1, col = "blue")  
        } else if (str_detect(Polymer[i,sp], "AF") == T){
          points(x = log10(NOEC.det[i,sp]),
                 y = prop[ind[sp]],
                 pch = 15, cex = 1, col = "mediumpurple")
        }
      }
      else if (str_detect((Shape[i,sp]), "irregular") && !is.na(Shape[i,sp])){
        if (str_detect(Polymer[i,sp], "PS") == T){
          points(x = log10(NOEC.det[i,sp]),
                 y = prop[ind[sp]],
                 pch = 17, cex = 1, col = "gray18")
        } else if (str_detect(Polymer[i,sp], "PET") == T){
          points(x = log10(NOEC.det[i,sp]),
                 y = prop[ind[sp]],
                 pch = 17, cex = 1, col = "green4") 
        } else if (str_detect(Polymer[i,sp], "PVC") == T){
          points(x = log10(NOEC.det[i,sp]),
                 y = prop[ind[sp]],
                 pch = 17, cex = 1, col = "skyblue")
        } else if (str_detect(Polymer[i,sp], "PE") == T){
          points(x = log10(NOEC.det[i,sp]),
                 y = prop[ind[sp]],
                 pch = 17, cex = 1, col = "darkgray")
        } else if (str_detect(Polymer[i,sp], "PLA") == T){
          points(x = log10(NOEC.det[i,sp]),
                 y = prop[ind[sp]],
                 pch = 17, cex = 1, col = "lightgreen")
        } else if (str_detect(Polymer[i,sp], "PA") == T){
          points(x = log10(NOEC.det[i,sp]),
                 y = prop[ind[sp]],
                 pch = 17, cex = 1, col = "red")
        } else if (str_detect(Polymer[i,sp], "PUR") == T){
          points(x = log10(NOEC.det[i,sp]),
                 y = prop[ind[sp]],
                 pch = 17, cex = 1, col = "yellow")
        } else if (str_detect(Polymer[i,sp], "PP") == T){
          points(x = log10(NOEC.det[i,sp]),
                 y = prop[ind[sp]],
                 pch = 17, cex = 1, col = "blue")
        } else if (str_detect(Polymer[i,sp], "AF") == T){
          points(x = log10(NOEC.det[i,sp]),
                 y = prop[ind[sp]],
                 pch = 17, cex = 1, col = "mediumpurple")
        }
      }
      
      xoffset <- c(0.5, 0.5, 0.5, 0.5, 0.5, 0.5, 0.5, 0.5, 0.5, 0.5, 0.5, 0.5, 0.5, 0.5, 
                   0.5, 0.5, 0.5, 0.5, 0.5, 0.5, 0.5, 0.5, 0.5, 0.5, 0.5, 0.5, 0.5, 0.5)
      
      text(x = min(log10((NOEC.det[,sp])), na.rm = T) - xoffset[sp],
           adj = 1,
           y = prop[ind[sp]],
           labels = colnames(Datapoints)[sp],
           cex = 0.6, font = 3)
    } 
  }
  
  
  #legend("bottomright", c("Mean PSSD", "PMMA", "PHB", "PS", "PS-COOH", "PS-NH2","with Sodium Azide"),
  #pch = c(NA, 16, 16, 16, 16, 16, 2), col = c("firebrick4", "yellow", "forestgreen", "black","red", "blue","black"),
  #lty = c(1, NA, NA, NA, NA, NA, NA), cex = 0.7, bg = "white", box.lwd = 0.4)
  
  ### NEW PLOT FOR THE PNEC ON NEW PDF PAGE
  
  # calculate it
  # activate when running for the first time
  PNEC1 <- apply(pSSD, 2, function(x) quantile(x, probs = 0.05, type = 1))
  hist(PNEC1, freq = F, breaks = 100, col = rgb(255, 127, 80, max = 255, alpha = 50), border = "coral",
       xlab = "PNEC (ug/L)", ylab = "Probability density", main = "Probability density of the PNEC", lwd = 0.5)
  #lines(density(log(PNEC,10)), lwd = 2, col = "red")
  box(lwd = 0.5)
  
  
  #activate when running for the first time
    save(PNEC1, file = "PNEC_mass.RData")
  
  #close the graph
  dev.off()
}

# PNEC value
Mode_Y	<- function(x)
{
  dens	<- density(x)
  ind		<- which(dens$y==max(dens$y))
  dens$x[ind]
}

Stat_PNEC <- matrix(NA, 2, 8)

Stat_PNEC[1,] <- c("Min", "Q5", "Q25", "Mean", "Mode", "Q75", "Q95", "Max")

Stat_PNEC[2,1] <- min(PNEC1)
Stat_PNEC[2,2] <- quantile(PNEC1, 0.05)
Stat_PNEC[2,3] <- quantile(PNEC1, 0.25)
Stat_PNEC[2,4] <- mean(PNEC1)
Stat_PNEC[2,5] <- Mode_Y(PNEC1)
Stat_PNEC[2,6] <- quantile(PNEC1, 0.75)
Stat_PNEC[2,7] <- quantile(PNEC1, 0.95)
Stat_PNEC[2,8] <- max(PNEC1)

write.csv(Stat_PNEC, file = "PSSD/updated/Stats_tier1_2_large.csv")

data.frame(t(Stat_PNEC))
```

## PSSD Ggplot code
### Preparation
```{r}
library(ggplot2)
library(dplyr)

# Assuming ECDF.data is already calculated as shown in your code
iv_vec <- seq(-3, 7, length.out = ncol(ECDF.data))

# Preparing dataframe for ggplot
ecdf_df <- data.frame(iv = rep(iv_vec, times = nrow(ECDF.data)), 
                      ecdf_value = as.vector(t(ECDF.data)))

# Calculating quantiles for shading
quantiles_df <- ecdf_df %>%
  group_by(iv) %>%
  summarise(min = min(ecdf_value),
            max = max(ecdf_value),
            q05 = quantile(ecdf_value, 0.05),
            q95 = quantile(ecdf_value, 0.95),
            q25 = quantile(ecdf_value, 0.25),
            q75 = quantile(ecdf_value, 0.75),
            mean = mean(ecdf_value)) %>%
  ungroup()

# deterministic values of NOEC (for plotting)
NOEC.det <- Datapoints /(UFdd * UFt)

 # calculate the geometric mean
  NOEC.gmean <- apply(NOEC.det, 2, function(x) geometric.mean(x, na.rm =T ))
  
  # calculate where to put the endpoints on the y axis
  #prop <- 1:ncol(Datapoints)/(ncol(Datapoints)+1)
  prop <- rep(NA, length(NOEC.gmean))
  for (i in 1:length(NOEC.gmean)) {
    prop[i] <- i/(length(NOEC.gmean))-(1/(length(NOEC.gmean))/2)
  }

  # Calculate where to put the endpoints on the y axis
prop <- seq_along(NOEC.gmean) / (length(NOEC.gmean) + 1)
  
ind <- rank(NOEC.gmean)

# Creating a dataframe for ggplot
species_data <- data.frame(
  Species = colnames(NOEC.det),
  NOEC_log10 = log10(NOEC.gmean),
  NOEC = NOEC.gmean,
  Prop = ind/(length(NOEC.gmean) + 1)
)

```

###Metadata
```{r}
# Join metadata to species
group <- tier1_2_large %>% 
  distinct(Group, Species)

# env
environment <- tier1_2_large %>% 
  distinct(Species, environment)

species_data <-species_data %>% 
  left_join(group, by = "Species") %>% 
  left_join(environment, by = "Species")


# Adjusting ind if necessary
# Assuming specific adjustments to 'prop' based on 'ind' calculations or other criteria
# Include those adjustments here if they are necessary for your plot
# Example:
# species_data$Prop[4] <- 4 / (length(NOEC.gmean) + 1)
# Adjust as needed based on your original 'ind' vector adjustments

```

## Plotting
```{r}
ggplot() +
  geom_ribbon(data = quantiles_df, aes(x = iv, ymin = min, ymax = max), fill = "coral", alpha = 0.1) +
  geom_ribbon(data = quantiles_df, aes(x = iv, ymin = q05, ymax = q95), fill = "coral", alpha = 0.3) +
  geom_ribbon(data = quantiles_df, aes(x = iv, ymin = q25, ymax = q75), fill = "coral", alpha = 0.8) +
  geom_point(data = species_data, aes(x = NOEC_log10, y = Prop, color = Group)) +
  geom_text(data = species_data, aes(x = NOEC_log10, y = Prop, label = Species, color = Group), 
            hjust = -0.2, vjust = 0,
            fontface = "italic",
            family = "helvetica"
            ) +
  geom_line(data = quantiles_df, aes(x = iv, y = mean), color = "firebrick4", size = 1) +
  scale_x_continuous(name = "NOEC (particles/L)", breaks = c(-4, -2, 0, 2, 4, 6, 8,10),
                     labels = expression(10^-4, 10^-2, 1, 10^2, 10^4, 10^6, 10^8, 10^10),
                     limits = c(-4, 10)) +
  scale_y_continuous(name = "Cumulative probability", limits = c(0, 1), breaks = seq(0, 1, 0.2)) +
  #theme_minimal() +
  dark_theme_minimal(base_size = 15) +
  ggtitle("Microplastics PSSD")

# Note: Replace `mean` with `median` in `aes(y = ...)` of `geom_line` for plotting the median

# Saving the plot to a PDF file
```

## PNEC
```{r}
library(ggplot2)

# Assuming pSSD and PNEC1 are already calculated as shown
# Calculate PNEC1 if not already calculated
# PNEC1 <- apply(pSSD, 2, function(x) quantile(x, probs = 0.05, type = 1))

# Convert PNEC1 to a dataframe for ggplot
PNEC1_df <- data.frame(PNEC1 = PNEC1)

# Plotting
ggplot(PNEC1_df, aes(x = PNEC1)) +
  geom_histogram(aes(y = ..density..), fill = rgb(255, 127, 80, max = 255, alpha = 50), color = "coral", bins = 100) +
  labs(x = "PNEC (Particles/L)", y = "Probability density", title = "Probability density of the PNEC") +
  #theme_minimal() +
  geom_density(col = "red", lwd = 1) +  # Optionally add a density line
  dark_theme_minimal(base_size = 18)

# Save PNEC1 if needed
# save(PNEC1, file = "PNEC_mass.RData")
```

```{r}
# Mode calculation using density
Mode_Y <- function(x) {
  dens <- density(x)
  ind <- which(dens$y == max(dens$y))
  return(dens$x[ind])
}

# Calculate statistics for PNEC1
Stat_PNEC <- data.frame(
  Min = min(PNEC1),
  Q5 = quantile(PNEC1, 0.05),
  Q25 = quantile(PNEC1, 0.25),
  Mean = mean(PNEC1),
  Mode = Mode_Y(PNEC1),
  Q75 = quantile(PNEC1, 0.75),
  Q95 = quantile(PNEC1, 0.95),
  Max = max(PNEC1)
)

# Transposing for similar structure to your matrix
Stat_PNEC_t <- t(Stat_PNEC)
colnames(Stat_PNEC_t) <- "Statistics"
print(Stat_PNEC_t)

```

