---
title: "PSSDplusplus walkthrough"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{PSSDplusplus walkthrough}
  %\VignetteEngine{knitr::rmarkdown}
  \usepackage[utf8]{inputenc}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(comment = "#>", collapse = TRUE, verbose = FALSE)

required_pkgs <- c(
  "sensobol",
  "truncnorm",
  "mc2d",
  "magrittr",
  "trapezoid",
  "ggpubr",
  "gtsummary",
  "doParallel",
  "doSNOW",
  "tictoc",
  "crayon",
  "Matrix",
  "reshape2",
  "future",
  "future.apply",
  "progressr"
)

missing <- required_pkgs[!vapply(required_pkgs, requireNamespace, logical(1), quietly = TRUE)]
if (length(missing)) {
  stop("Missing required packages for this vignette: ",
       paste(missing, collapse = ", "))
}
library(PSSDplusplus)
library(trapezoid)
`%>%` <- magrittr::`%>%`

# Automatically detect and set parallel processing options based on rendering or running locally
parallel_enabled <- TRUE
if (knitr::is_html_output() || knitr::is_latex_output()) {
  parallel_enabled <- FALSE
  future::plan(future::sequential)
}

workers <- if (parallel_enabled) max(1, parallel::detectCores() - 1) else 1
num_cores <- if (parallel_enabled) "auto" else 1
```

# Overview
This document provides a minimal, end-to-end walkthrough of the **PSSD++ probabilistic ecotoxicological threshold framework** implemented in the `PSSDplusplus` R package. The workflow demonstrates how microplastic (MP) hazard thresholds are derived across environments and Effect–Response Metrics (ERMs) using Monte Carlo uncertainty propagation and probabilistic Species Sensitivity Distributions (pSSDs).

The analytical pipeline mirrors that described in **Coffin et al. (2026)** and related manuscripts, while remaining compact enough for method development, debugging, and demonstration.

---

## Overview of the PSSD++ Framework

The PSSD++ framework integrates the following components:

- **Monte Carlo (MC) uncertainty propagation**
- **Probabilistic Species Sensitivity Distributions (pSSDs)**
- **Alignment of laboratory toxicity data to environmentally relevant MP particle distributions**
- **Parallel computation with deterministic caching**

The workflow proceeds through four main stages:

1. Preparation and filtering of toxicity inputs  
2. Sampling of alignment and uncertainty parameters  
3. Monte Carlo alignment of toxicity data  
4. Fitting and summarization of PSSD++ models  

```{r render-mode, include = FALSE}
render_mode <- knitr::is_html_output() || knitr::is_latex_output()
if (render_mode) {
  knitr::opts_chunk$set(cache = FALSE)
  options(progressr.enable = FALSE)
}

parallel_enabled <- !render_mode
if (!parallel_enabled) {
  future::plan(future::sequential)
}

workers <- if (parallel_enabled) max(1, parallel::detectCores() - 1) else 1
num_cores <- if (parallel_enabled) "auto" else 1
base_cache_dir <- file.path(tempdir(), "pssd_cache")
base_output_path <- file.path(tempdir(), "pssd_figures")
```

## 1) Prepare inputs
By default, the package uses a preprocessed **ToMEx 2.0–derived ecotoxicity dataset**, distributed with the package.
```{r}
# load in toxicity data (ToMEx 2.0 is default, and is supplied with package)
tox_data <- tomex2
head(tox_data) # view the first few rows of the toxicity data
```

The user may choose to filter the toxicity data based on specific criteria (e.g., species, exposure duration, endpoints, quality criteria, etc.). Each of these criteria can be applied using the `filter()` function from the `dplyr` package. Filtering options applied in Mehinto et al. (2022) are applied below:
```{r}
tox_data <- tox_data |> 
  dplyr::filter(
    Group != "Bacterium", # Drop studies with Bacterium group
    Group != "Plant", # Drop studies with Plant group
    effect.metric != "HONEC", # Drop studies with HONEC effect metric
    tier_zero_tech_f == "Red Criteria Passed", #minimum QC for technical criteria
    tier_zero_risk_f == "Red Criteria Passed", #minimum QC for risk criteria
    risk.13 != 0 #Drop studies that received a score of 0 for endpoints criteria (this also drops studies that have not yet been)
  )
```

## 2) Sample alignment parameters
Use the defaults from the manuscript (i.e., Kooi et al. 2021)[https://doi.org/10.1016/j.envpol.2021.117898] or pass a custom `params` table.
Note that the parameter names cannot change - regardless of what the user's specified environment is. For instance, if the user would like to model a brackish water environment, simply update the parameters for either 'marine' or 'freshwater' with the brackish parameters, while keeping the original names of parameters intact.

```{r}
param_default_values <- PSSDplusplus::param_default_values
#view parameters
head(param_default_values)
```

A parameter distribution dataset may be simulated using each respective parameters' distributions (e.g., alpha power law, normal distributions, etc.). By default, the package uses a Latin Hypercube Sampling approach to generate the parameter matrix. If the user wants to user alternative distributions for each parameter, the matrix_function() should be modified accordingly. This version supports surface water and sediment parameter draws. Use the include_* toggles to control which compartments are included in the LHS matrix (and therefore the matrix size).
```{r}
param_matrix <- matrix_function(
  n_sobol = 5, #size of matrix (10,000 is recommended)
  params = param_default_values, #parameter distribution values -replace with user's specifications
  upper.tissue.truncation.limit = 500, # maximum particle size (microns) to truncate tissue tranlocation size limit distribution to based on biological plausibility
  x1M_set = 1, #minimum particle size to generate distributions to in microns
  x2D_set = 5000, #maximum particle size to generate distributions to in microns
  include_marine_surface_water = TRUE,
  include_freshwater_surface_water = TRUE,
  include_marine_sediment = TRUE,
  include_freshwater_sediment = TRUE
)
```

Visualize the sampled draws (each element is a ggplot object):

```{r parameter_histogram_plots}
param_plots <- parameter_histograms_function(param_matrix) # generate parameter distribution plots
#visualize plots
param_plots$alpha_combined_plot # distribution of alpha parameters 
param_plots$tissue_body #distribution of tissue translocation and mouth size opening model parameters
```

To focus on sediment-only draws, pass sediment compartments explicitly:

```{r sediment_parameter_histograms}
sediment_param_plots <- parameter_histograms_function(
  param_matrix,
  compartments = c("Marine Sediment", "Freshwater Sediment")
)
sediment_param_plots$alpha_combined_plot
```

## 3) Run Monte Carlo alignments
The following code aligns the toxicity data to the parameter matrix using Monte Carlo simulations. 
```{r MC-sim alignments}
message("Starting MC_sim_align_parallel...")
MC_sim_df <- MC_sim_align_parallel(
  tox_data = tox_data, # toxicity data (ToMEx 2.0 is default)
  param_matrix = param_matrix, # parameter matrix generated in prior cell
  n_sim = 3, # size of parameter matrix (ensure this is the same as the size of the matrix generated)
  x1D_set = 1, # minimum particle size to generate distributions to (in microns)
  x2D_set = 5000, # maximum particle size to generate distributions to (in microns)
  num_cores = num_cores # number of cores to use for parallel processing (specify )
)
message("Finished MC_sim_align_parallel.")
```

### Split the aligned data into ERM- and environment-specific subsets
Split the aligned data into ERM- and environment-specific registries to replicate the results from Coffin et al. (2026). Surface water and sediment use different dose units, so build separate registries for each. Surface water uses `dose_new_particles_L`, while sediment uses `dose_new_particles_kg`.
```{r}
# ---- Surface water (particles/L) ----
results_df_food_water <- dplyr::filter(
  MC_sim_df,
  environment %in% c("Freshwater", "Marine"),
  ingestible != "not ingestible",
  particles_L_food_dilution > 0,
  Group != "Algae"
) |>
  dplyr::mutate(dose_new_particles_L = particles_L_food_dilution) |>
  tidyr::drop_na(particles_L_food_dilution)
results_df_food_water_t3_t4 <- dplyr::filter(
  results_df_food_water,
  risk.13 != 1,
  bio_f %in% c("Organism", "Population")
)

results_df_tissue_water <- dplyr::filter(
  MC_sim_df,
  environment %in% c("Freshwater", "Marine"),
  translocatable != "not translocatable",
  particles_L_ox_stress > 0
) |>
  dplyr::mutate(dose_new_particles_L = particles_L_ox_stress) |>
  tidyr::drop_na(particles_L_ox_stress)
results_df_tissue_water_t3_t4 <- dplyr::filter(
  results_df_tissue_water,
  risk.13 != 1,
  bio_f %in% c("Organism", "Population")
)

erm_registry_water <- list(
  "Food Dilution" = list(base = results_df_food_water, t3_t4 = results_df_food_water_t3_t4),
  "Tissue Translocation" = list(base = results_df_tissue_water, t3_t4 = results_df_tissue_water_t3_t4)
)

# ---- Sediment (particles/kg) ----
results_df_food_sediment <- dplyr::filter(
  MC_sim_df,
  environment %in% c("Freshwater Sediment", "Marine Sediment"),
  ingestible != "not ingestible",
  particles_kg_food_dilution > 0,
  Group != "Algae"
) |>
  dplyr::mutate(dose_new_particles_kg = particles_kg_food_dilution) |>
  tidyr::drop_na(particles_kg_food_dilution)
results_df_food_sediment_t3_t4 <- dplyr::filter(
  results_df_food_sediment,
  risk.13 != 1,
  bio_f %in% c("Organism", "Population")
)

results_df_tissue_sediment <- dplyr::filter(
  MC_sim_df,
  environment %in% c("Freshwater Sediment", "Marine Sediment"),
  translocatable != "not translocatable",
  particles_kg_ox_stress > 0
) |>
  dplyr::mutate(dose_new_particles_kg = particles_kg_ox_stress) |>
  tidyr::drop_na(particles_kg_ox_stress)
results_df_tissue_sediment_t3_t4 <- dplyr::filter(
  results_df_tissue_sediment,
  risk.13 != 1,
  bio_f %in% c("Organism", "Population")
)

erm_registry_sediment <- list(
  "Food Dilution" = list(base = results_df_food_sediment, t3_t4 = results_df_food_sediment_t3_t4),
  "Tissue Translocation" = list(base = results_df_tissue_sediment, t3_t4 = results_df_tissue_sediment_t3_t4)
)
```

## 4) Fit PSSD++s
The following code performs the fitting of PSSD++ models to the aligned data for surface water and sediment separately:
```{r}
message("Starting make_all_pSSDs...")
# call object for usage by rtrunc later (this ONLY needs to be done in the vignette, which cannot load other libraries in the file)
rtrapezoid <- trapezoid::rtrapezoid
ptrapezoid <- trapezoid::ptrapezoid
qtrapezoid <- trapezoid::qtrapezoid
dtrapezoid <- trapezoid::dtrapezoid

# make all pSSDs for surface water
pSSDs_water <- make_all_pSSDs(
  MC_sim_df = MC_sim_df, # ensure aligned data is included
  environments = c("Freshwater", "Marine"), #specify which environments to include
  erm_registry = erm_registry_water,
  sim = 3, # specify the number of simulations to run
  cv_uf = 0.5, # specify the coefficient of variation for uncertainty factors (acute -> chronic and LOEC/EC50 -> NOEC)
  rmore_method = "step", # method to handle pSSD distribution building (options = 'step' i.e., original trapezoidal method in Wigger et al., 2020, or 'lognormal' - shortcut developed in Coffin et al. (2025))
  quantile_type = 8, # quantile type for distribution fitting (8 is default, and use of others may cause unexpected or unreliable results)
  parallel = parallel_enabled, #whether or not to use parallel processing
  workers = workers, # number of worker processes to use for parallel processing
  base_cache_dir = file.path(tempdir(), "pssd_cache"), # directory to store cached pSSD objects (important when running high number of simulations to resume if needed)
  base_output_path = file.path(tempdir(), "pssd_figures"), # directory to store output figures
  overwrite_cache = TRUE, # whether or not to overwrite cached pSSD objects
  debug = T # whether or not to enable debug mode
)

# make all pSSDs for sediment (kg-based doses)
  pSSDs_sediment <- make_all_pSSDs(
    MC_sim_df = MC_sim_df,
    environments = c("Freshwater Sediment", "Marine Sediment"),
    erm_registry = erm_registry_sediment,
    sim = 3,
    cv_uf = 0.5,
    rmore_method = "step",
    quantile_type = 8,
    parallel = parallel_enabled,
    workers = workers,
    base_cache_dir = file.path(tempdir(), "pssd_cache_sediment"),
    base_output_path = file.path(tempdir(), "pssd_figures_sediment"),
    overwrite_cache = TRUE,
    debug = F
  )
```


### Examine pSSD++ output
Each ERM x Environment combination is saved as a list element, with corresponding PNEC quantiles, summary statistics, and plots.

#### PNEC Summary
By default the pSSD++ performs a quantile regression to estimate PNEC values for Tiers 3 and 4 of the Mehinto et al. (2022) risk assessment framework.
```{r, eval = FALSE}
PNEC_summary_water <- summarize_PNECs(pSSDs_water)
head(PNEC_summary_water)

PNEC_summary_sediment <- summarize_PNECs(pSSDs_sediment)
head(PNEC_summary_sediment)
```

PNEC distributions are visualized using ggplot:
```{r}
pSSDs_water$`Tier3_Freshwater_Food Dilution`$PNEC_plot_05
```

#### pSSD++ Plots
The pSSD++ plots show the relationship between the exposure concentration and the predicted effect levels.
```{r}
# view pSSD plot for example:
pSSDs_water$`Tier3_Freshwater_Food Dilution`$pSSD_plot
```

```{r}
pSSDs_water$`Tier3_Marine_Tissue Translocation`$pSSD_plot
```
#### Arranged plot
A manuscript-ready pSSD++ with PNEC distribution plots for Tiers 3 and 4 are generated.
```{r}
pSSDs_water$`Tier3_Freshwater_Food Dilution`$arranged_plot
```

Sediment plots use the same access pattern, but with sediment environments:

```{r}
pSSDs_sediment$`Tier3_Freshwater Sediment_Food Dilution`$pSSD_plot
```

The cache and figure outputs default to temporary directories; override the paths above to persist artifacts between sessions.

### Combining environments
In some cases, one may want to combine pSSD outputs across different environments (e.g., freshwater + marine sediment due to low number of species and/or mixed species environment such as an estuary). This can be done accomplished by overwriting the environment variable in the toxicity dataset before running the pSSD workflow. An example is provided below.

```{r}
# overwrite environment variable
tox_data_combined <- tox_data |> dplyr::mutate(env_f = "Freshwater")

#MC-Align
MC_sim_df_combined <- MC_sim_align_parallel(
  tox_data = tox_data_combined, # toxicity data (ToMEx 2.0 is default)
  param_matrix = param_matrix, # parameter matrix generated in prior cell
  n_sim = 3, # size of parameter matrix (ensure this is the same as the size of the matrix generated)
  x1D_set = 1, # minimum particle size to generate distributions to (in microns)
  x2D_set = 5000, # maximum particle size to generate distributions to (in microns)
  num_cores = num_cores # number of cores to use for parallel processing (specify )
)

# Filter results to valid data
results_df_food_sediment_combined <- dplyr::filter(
  MC_sim_df_combined,
  ingestible != "not ingestible",
  particles_kg_food_dilution > 0,
  Group != "Algae") |>
  dplyr::mutate(dose_new_particles_kg = particles_kg_food_dilution) |>
  tidyr::drop_na(particles_kg_food_dilution)

# build ERM registry for pSSD (just one tier for food dilution for minimal reprex)
erm_registry_sediment_combined <- list(
  "Food Dilution" = list(base = results_df_food_sediment_combined, t3_t4 = results_df_food_sediment_combined),
  "Tissue Translocation" = list(base = NA, t3_t4 = NA)
)

# run pSSD workflow
pSSDs_sediment_combined <- make_all_pSSDs(
  MC_sim_df = MC_sim_df_combined, # ensure aligned data is included
  environments = c("Freshwater Sediment"), #marine and freshwater combined to fresdhwater
  erm_registry = erm_registry_sediment_combined,
  sim = 3, 
  base_cache_dir = file.path(tempdir(), "pssd_cache_combined"), # directory to store cached pSSD objects (important when running high number of simulations to resume if needed)
  base_output_path = file.path(tempdir(), "pssd_figures_combined"), # directory to store output figures
  overwrite_cache = TRUE # whether or not to overwrite cached pSSD objects
)

# inspect results
pSSDs_sediment_combined$`Tier3_Freshwater Sediment_Food Dilution`$pSSD_plot
```
