---
title: "PSSDplusplus walkthrough"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{PSSDplusplus walkthrough}
  %\VignetteEngine{knitr::rmarkdown}
  \usepackage[utf8]{inputenc}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(comment = "#>", collapse = TRUE, verbose = FALSE)

# sensobol powers matrix_function()'s Sobol'/LHS sampling and is required for
# this vignette's Monte Carlo workflow. All other optional packages used below
# (ggpubr, cols4all, ggtext, doSNOW/foreach, future/future.apply, progressr,
# tictoc, crayon) degrade gracefully when missing, so they are not required here.
if (!requireNamespace("sensobol", quietly = TRUE)) {
  stop("Missing required package for this vignette: sensobol")
}
has_plotly <- requireNamespace("plotly", quietly = TRUE)
library(PSSDplusplus)
library(trapezoid)
`%>%` <- magrittr::`%>%`

# Default to parallel when run interactively; overridden to FALSE when rendering.
parallel_enabled <- TRUE

render_mode <- knitr::is_html_output() || knitr::is_latex_output()
if (render_mode) {
  knitr::opts_chunk$set(cache = FALSE)
  options(progressr.enable = FALSE)
  parallel_enabled <- FALSE
  future::plan(future::sequential)
}

# Useful parallelism is bounded by the job count, not the core count:
#   MC_sim_align_parallel: one PSOCK job per n_sim row. PSOCK startup + data
#     serialisation dominates for n_sim <= 20, so the function auto-falls back
#     to sequential in that range. For n_sim 1000+, 4-8 cores is the sweet spot.
#   make_all_pSSDs: one future job per tier x environment x ERM combination.
#     Default (Tier3 x 2 envs x 2 ERMs) = 4 combos; the function internally
#     caps workers to nrow(combo_tbl), so extra workers are never launched.
# Setting workers = detectCores()-1 on a 16-core machine wastes 12 idle sessions.
workers <- if (parallel_enabled) min(max(1L, parallel::detectCores() - 1L), 4L) else 1
num_cores <- if (parallel_enabled) "auto" else 1
base_cache_dir <- file.path(tempdir(), "pssd_cache")
base_output_path <- file.path(tempdir(), "pssd_figures")
```

# Overview
This document provides a minimal, end-to-end walkthrough of the **PSSD++ probabilistic ecotoxicological threshold framework** implemented in the `PSSDplusplus` R package. The workflow demonstrates how microplastic (MP) hazard thresholds are derived across environments and Effect–Response Metrics (ERMs) using Monte Carlo uncertainty propagation and probabilistic Species Sensitivity Distributions (pSSDs).

The analytical pipeline mirrors that described in **Coffin et al. (2026)** and related manuscripts, while remaining compact enough for method development, debugging, and demonstration.

---

## Overview of the PSSD++ Framework

The PSSD++ framework integrates the following components:

- **Monte Carlo (MC) uncertainty propagation**
- **Probabilistic Species Sensitivity Distributions (pSSDs)**
- **Alignment of laboratory toxicity data to environmentally relevant MP particle distributions**
- **Parallel computation with deterministic caching**

The workflow proceeds through four main stages:

1. Preparation and filtering of toxicity inputs  
2. Sampling of alignment and uncertainty parameters  
3. Monte Carlo alignment of toxicity data  
4. Fitting and summarization of PSSD++ models  


## Quick start: alignment → traditional SSDs

This section demonstrates the complete basic workflow in a few steps using the
default (deterministic) parameter values and the bundled `tomex2` dataset.
Use this path when you want single-run, non-Monte-Carlo SSDs — for example,
to explore the data, reproduce Tier 1–4 thresholds at default parameters, or
validate results against the full probabilistic workflow.

### Step A — Filter and align

```{r basic-align, message = FALSE}
library(PSSDplusplus)
library(dplyr)

tox_data <- tomex2_mini #beta-testing

# 1. Load and (optionally) filter ToMEx 2.0 
tox_filtered <- tox_data |>
   # demonstrating filters used for threshold development. Adjust as needed.
  dplyr::filter(
    Group != "Bacterium",
    Group != "Plant",
    effect.metric != "HONEC",
    tier_zero_tech_f == "Red Criteria Passed",
    tier_zero_risk_f == "Red Criteria Passed",
    risk.13 != 0
  )

# 2. Align to ERMs using default parameters (single deterministic run)
aligned <- align_data(tox_filtered)
```

`align_data()` adds columns such as `EC_env_sa.particles.mL_trans` (tissue
translocation, particles/mL) and `EC_env_v.particles.mL_ingest` (food dilution,
particles/mL) to every row. Sediment equivalents use `particles.kg` units.

### Step B — Prepare data for SSD fitting

After alignment the dose column must be named `dose_new`. The snippet below
prepares a food-dilution freshwater surface-water subset for Tiers 1–4:

`make_tiered_SSDs()` uses `dose_new` as provided and does not apply assessment
factors internally. The example divides by `af.time * af.noec` to recreate the
Mehinto et al. threshold preparation, where `af.time` converts acute values to
chronic equivalents and `af.noec` converts LOEC/ECx/LCx values to
NOEC-equivalent values. Omit this division when inputs are already
NOEC-equivalent or when running an unadjusted sensitivity analysis.

```{r basic-prep}
# Surface-water food dilution subset (particles/L = particles/mL * 1000)
fd_water <- aligned |>
  dplyr::filter(
    environment %in% c("Freshwater", "Marine"),
    ingestible != "not ingestible",
    EC_env_v.particles.mL_ingest > 0,
    Group != "Algae",
    shape_f != "Not Reported",
    poly_f  != "Not Reported"
  ) |>
  dplyr::mutate(
    dose_new = (EC_env_v.particles.mL_ingest / (af.time * af.noec)) * 1000
  ) |>
  tidyr::drop_na(dose_new) |>
  dplyr::filter(dose_new > 0)
```

### Step C — Fit tiered SSDs with `make_tiered_SSDs()`

The tiers follow the risk assessment framework defined in
[Mehinto et al. (2022)](https://link.springer.com/article/10.1186/s43591-022-00033-3),
which grades data quality and protectiveness from Tier 1 (all endpoints, most
protective) through Tier 4 (organism/population endpoints only, predictive):

| `tier` | Data filter | Collapse | HCx |
|--------|-------------|----------|-----|
| 1 | All endpoints | 25th percentile | HC5 lower confidence limit |
| 2 | All endpoints | 25th percentile | HC5 estimate + 95% CI |
| 3 | Organism/Population, `risk.13 ≠ 1` | Median | HC5 estimate + 95% CI |
| 4 | Organism/Population, `risk.13 ≠ 1` | Median | HC10 estimate + 95% CI |

`make_tiered_SSDs()` is the single entry point for all four tiers:

```{r basic-ssd}
set.seed(42)

# Tier 1 — most protective (Mehinto et al. 2022); returns HC5 lower CL
result_t1 <- make_tiered_SSDs(fd_water, tier = 1)

# Tier 2 — HC5 estimate + 95% CI
result_t2 <- make_tiered_SSDs(fd_water, tier = 2)

# Tier 3 — organism/population only, median collapse, HC5 + CI
result_t3 <- make_tiered_SSDs(fd_water, tier = 3)

# Tier 4 — same filter as Tier 3, HC10 + CI
result_t4 <- make_tiered_SSDs(fd_water, tier = 4)

# Inspect:
result_t1$hc5lcl  # particles/L
result_t2$hcx_est
```

For the full Monte Carlo workflow (propagating alignment-parameter uncertainty
across hundreds of simulations), continue to Section 1 below.

---

### Step D — Plot the SSD

`make_tiered_SSDs()` returns a `tiered_SSD` S3 object that bundles everything
needed to reproduce the SSD plot.

#### Static plot with `autoplot()`

Call `ggplot2::autoplot()` on the result. Dashed crosshair lines intersect at
the actual HCx position on the curve. Adjust `base_size`, `point_size`, and
`label_size` to scale the plot for different figure dimensions:

```{r basic-plot, fig.width = 7, fig.height = 5}
# Print a compact summary
print(result_t3)

# Publication-ready static plot
ggplot2::autoplot(result_t3,
                  x_label    = "particles/L",
                  erm_label  = "Food Dilution",
                  base_size  = 12,    # overall font size
                  point_size = 2,     # species data points
                  label_size = 3.5)   # repelled species labels
```

#### Interactive plot with `plotly_ssd()`

`plotly_ssd()` produces the same layout as `autoplot()` but as an interactive
[plotly](https://plotly.com/r/) widget. Requires the `plotly` package
(`install.packages("plotly")`). The arrow annotation points directly to the HCx
position on the SSD curve:

```{r plotly-plot, eval = has_plotly, fig.width = 7, fig.height = 5}
# Interactive SSD with hover tooltips and HCx arrow annotation
plotly_ssd(result_t3,
           x_label    = "particles/L",
           erm_label  = "Food Dilution",
           lower_size = 1,      # optional: particle size range for axis title
           upper_size = 5000)
```

```{r plotly-plot-unavailable, eval = !has_plotly, echo = FALSE, results = "asis"}
knitr::asis_output(
  "> The interactive Plotly widget is skipped because the optional `plotly` package is not installed. Install it with `install.packages(\"plotly\")` and re-knit this vignette to render the widget.\n"
)
```

The `tiered_SSD` object also exposes the underlying data for fully custom plots:

```{r basic-plot-components, eval = FALSE}
# Model-averaged SSD predictions with CI (aoc_pred in Shiny)
result_t3$predictions    # columns: proportion, est, lcl, ucl, ...

# Species-level collapsed data with empirical CDF fractions (aoc_ssd)
result_t3$collapsed_data # columns: Species, Group, Conc, frac

# Single-row HCx annotation data with est_format (aoc_hc)
result_t3$hc_data

# The raw ssdtools fit object for re-fitting or custom predictions
result_t3$fit_dists
```

---

## 1) Prepare inputs
This vignette uses **`tomex2_mini`**, a compact 10-species (5 freshwater + 5 marine) subset
of the full ToMEx 2.0 dataset bundled with the package. It is pre-filtered to pass all Tier 1–4
quality criteria and runs end-to-end in seconds, making it ideal for exploration and testing.
To replicate the full analysis from Coffin et al. (2026), swap `tomex2_mini` for `tomex2`.
```{r}
# load in toxicity data (tomex2_mini is a compact 10-species subset of ToMEx 2.0. that is pre-filtered for quick testing.)
tox_data <- tomex2_mini #beta-testing
#tox_data <- tomex2 # full dataset
head(tox_data) # view the first few rows of the toxicity data
```

Users working with the full `tomex2` dataset may filter by specific criteria using `dplyr::filter()`.
The filters below (applied in Mehinto et al. 2022) are already satisfied in `tomex2_mini`, so they
are shown here as documentation of the filtering logic rather than an active transformation:
```{r}
tox_data <- tox_data |>
  dplyr::filter(
    Group != "Bacterium", # Drop studies with Bacterium group
    Group != "Plant", # Drop studies with Plant group
    effect.metric != "HONEC", # Drop studies with HONEC effect metric
    tier_zero_tech_f == "Red Criteria Passed", #minimum QC for technical criteria
    tier_zero_risk_f == "Red Criteria Passed", #minimum QC for risk criteria
    risk.13 != 0 #Drop studies that received a score of 0 for endpoints criteria (this also drops studies that have not yet been)
  )
```

## 2) Sample alignment parameters
Use the defaults from the manuscript (i.e., Kooi et al. 2021)[https://doi.org/10.1016/j.envpol.2021.117898] or pass a custom `params` table.
Note that the parameter names cannot change - regardless of what the user's specified environment is. For instance, if the user would like to model a brackish water environment, simply update the parameters for either 'marine' or 'freshwater' with the brackish parameters, while keeping the original names of parameters intact.

```{r}
param_default_values <- PSSDplusplus::param_default_values
#view parameters
head(param_default_values)
```

A parameter distribution dataset may be simulated using each respective parameters' distributions (e.g., alpha power law, normal distributions, etc.). By default, the package uses a Latin Hypercube Sampling approach to generate the parameter matrix. If the user wants to user alternative distributions for each parameter, the matrix_function() should be modified accordingly. This version supports surface water and sediment parameter draws. Use the include_* toggles to control which compartments are included in the LHS matrix (and therefore the matrix size).
```{r}
param_matrix <- matrix_function(
  n_sobol = 5, #size of matrix (10,000 is recommended)
  params = param_default_values, #parameter distribution values -replace with user's specifications
  upper.tissue.truncation.limit = 500, # maximum particle size (microns) to truncate tissue tranlocation size limit distribution to based on biological plausibility
  x1M_set = 1, #minimum particle size to generate distributions to in microns
  x2D_set = 5000, #maximum particle size to generate distributions to in microns
  include_marine_surface_water = TRUE,
  include_freshwater_surface_water = TRUE,
  include_marine_sediment = TRUE,
  include_freshwater_sediment = TRUE
)
```

Visualize the sampled draws (each element is a ggplot object):

```{r parameter_histogram_plots}
param_plots <- parameter_histograms_function(param_matrix) # generate parameter distribution plots
#visualize plots
param_plots$alpha_combined_plot # distribution of alpha parameters 
param_plots$tissue_body #distribution of tissue translocation and mouth size opening model parameters
```

To focus on sediment-only draws, pass sediment compartments explicitly:

```{r sediment_parameter_histograms}
sediment_param_plots <- parameter_histograms_function(
  param_matrix,
  compartments = c("Marine Sediment", "Freshwater Sediment")
)
sediment_param_plots$alpha_combined_plot
```

## 3) Run Monte Carlo alignments
The following code aligns the toxicity data to the parameter matrix using Monte Carlo simulations. 
```{r MC-sim alignments}
message("Starting MC_sim_align_parallel...")
MC_sim_df <- MC_sim_align_parallel(
  tox_data = tox_data, # toxicity data (ToMEx 2.0 is default)
  param_matrix = param_matrix, # parameter matrix generated in prior cell
  n_sim = 10, # number of MC alignment draws per species (must not exceed nrow(param_matrix))
  x1D_set = 1, # minimum particle size to generate distributions to (in microns)
  x2D_set = 5000, # maximum particle size to generate distributions to (in microns)
  num_cores = num_cores # number of cores to use for parallel processing (specify )
)
message("Finished MC_sim_align_parallel.")
```

### Split the aligned data into ERM- and environment-specific subsets
Split the aligned data into ERM- and environment-specific registries to replicate the results from Coffin et al. (2026). Surface water and sediment use different dose units, so build separate registries for each. Surface water uses `dose_new_particles_L`, while sediment uses `dose_new_particles_kg`.
```{r}
# ---- Surface water (particles/L) ----
results_df_food_water <- dplyr::filter(
  MC_sim_df,
  environment %in% c("Freshwater", "Marine"),
  ingestible != "not ingestible",
  particles_L_food_dilution > 0,
  Group != "Algae"
) |>
  dplyr::mutate(dose_new_particles_L = particles_L_food_dilution) |>
  tidyr::drop_na(particles_L_food_dilution)
results_df_food_water_t3_t4 <- dplyr::filter(
  results_df_food_water,
  risk.13 != 1,
  bio_f %in% c("Organism", "Population")
)

results_df_tissue_water <- dplyr::filter(
  MC_sim_df,
  environment %in% c("Freshwater", "Marine"),
  translocatable != "not translocatable",
  particles_L_ox_stress > 0
) |>
  dplyr::mutate(dose_new_particles_L = particles_L_ox_stress) |>
  tidyr::drop_na(particles_L_ox_stress)
results_df_tissue_water_t3_t4 <- dplyr::filter(
  results_df_tissue_water,
  risk.13 != 1,
  bio_f %in% c("Organism", "Population")
)

erm_registry_water <- list(
  "Food Dilution" = list(base = results_df_food_water, t3_t4 = results_df_food_water_t3_t4),
  "Tissue Translocation" = list(base = results_df_tissue_water, t3_t4 = results_df_tissue_water_t3_t4)
)

# ---- Sediment (particles/kg) ----
results_df_food_sediment <- dplyr::filter(
  MC_sim_df,
  environment %in% c("Freshwater Sediment", "Marine Sediment"),
  ingestible != "not ingestible",
  particles_kg_food_dilution > 0,
  Group != "Algae"
) |>
  dplyr::mutate(dose_new_particles_kg = particles_kg_food_dilution) |>
  tidyr::drop_na(particles_kg_food_dilution)
results_df_food_sediment_t3_t4 <- dplyr::filter(
  results_df_food_sediment,
  risk.13 != 1,
  bio_f %in% c("Organism", "Population")
)

results_df_tissue_sediment <- dplyr::filter(
  MC_sim_df,
  environment %in% c("Freshwater Sediment", "Marine Sediment"),
  translocatable != "not translocatable",
  particles_kg_ox_stress > 0
) |>
  dplyr::mutate(dose_new_particles_kg = particles_kg_ox_stress) |>
  tidyr::drop_na(particles_kg_ox_stress)
results_df_tissue_sediment_t3_t4 <- dplyr::filter(
  results_df_tissue_sediment,
  risk.13 != 1,
  bio_f %in% c("Organism", "Population")
)

erm_registry_sediment <- list(
  "Food Dilution" = list(base = results_df_food_sediment, t3_t4 = results_df_food_sediment_t3_t4),
  "Tissue Translocation" = list(base = results_df_tissue_sediment, t3_t4 = results_df_tissue_sediment_t3_t4)
)
```

## 4) Fit PSSD++s
The following code performs the fitting of PSSD++ models to the aligned data for surface water and sediment separately:
```{r}
message("Starting make_all_pSSDs...")
# call object for usage by rtrunc later (this ONLY needs to be done in the vignette, which cannot load other libraries in the file)
rtrapezoid <- trapezoid::rtrapezoid
ptrapezoid <- trapezoid::ptrapezoid
qtrapezoid <- trapezoid::qtrapezoid
dtrapezoid <- trapezoid::dtrapezoid

message("Building surface water pSSDs...")
# make all pSSDs for surface water
pSSDs_water <- make_all_pSSDs(
  MC_sim_df = MC_sim_df, # ensure aligned data is included
  environments = c("Freshwater", "Marine"), #specify which environments to include
  erm_registry = erm_registry_water,
  sim = 3, # specify the number of simulations to run
  cv_uf = 0.5, # specify the coefficient of variation for uncertainty factors (acute -> chronic and LOEC/EC50 -> NOEC)
  apply_assessment_factors = TRUE, # set FALSE only for NOEC-equivalent inputs or unadjusted sensitivity checks
  rmore_method = "step", # method to handle pSSD distribution building (options = 'step' i.e., original trapezoidal method in Wigger et al., 2020, or 'lognormal' - shortcut developed in Coffin et al. (2025))
  quantile_type = 8, # quantile type for distribution fitting (8 is default, and use of others may cause unexpected or unreliable results)
  parallel = parallel_enabled, #whether or not to use parallel processing
  workers = workers, # number of worker processes to use for parallel processing
  base_cache_dir = file.path(tempdir(), "pssd_cache"), # directory to store cached pSSD objects (important when running high number of simulations to resume if needed)
  base_output_path = file.path(tempdir(), "pssd_figures"), # directory to store output figures
  overwrite_cache = TRUE, # whether or not to overwrite cached pSSD objects
  debug = F # T or F: whether or not to enable debug mode. Off by default.
)

message("Building sediment pSSDs...")
# make all pSSDs for sediment (kg-based doses)
  pSSDs_sediment <- make_all_pSSDs(
    MC_sim_df = MC_sim_df,
    environments = c("Freshwater Sediment", "Marine Sediment"),
    erm_registry = erm_registry_sediment,
    sim = 3,
    cv_uf = 0.5,
    apply_assessment_factors = TRUE,
    rmore_method = "step",
    quantile_type = 8,
    parallel = parallel_enabled,
    workers = workers,
    base_cache_dir = file.path(tempdir(), "pssd_cache_sediment"),
    base_output_path = file.path(tempdir(), "pssd_figures_sediment"),
    overwrite_cache = TRUE,
    debug = F
  )
```

By default, the PSSD++ workflow applies ToMEx assessment factors through
`af.time` and `af.noec`. `make_all_pSSDs()`, `run_pssd_reprex()`,
`do.pSSD_mod()`, and `do.pSSD()` accept `apply_assessment_factors = FALSE` to
treat those factors as 1 for sensitivity checks or data already expressed as
NOEC-equivalent values. In that mode the `UFt` and `UFdd` matrices are optional.


### Examine pSSD++ output
Each ERM x Environment combination is saved as a list element, with corresponding PNEC quantiles, summary statistics, and plots.

#### PNEC Summary
By default the pSSD++ performs a quantile regression to estimate PNEC values for Tiers 3 and 4 of the Mehinto et al. (2022) risk assessment framework.
```{r, eval = FALSE}
PNEC_summary_water <- summarize_PNECs(pSSDs_water)
head(PNEC_summary_water)

PNEC_summary_sediment <- summarize_PNECs(pSSDs_sediment)
head(PNEC_summary_sediment)
```

PNEC distributions are visualized using ggplot:
```{r}
pSSDs_water$`Tier3_Freshwater_Food Dilution`$PNEC_plot_05
```

#### pSSD++ Plots
The pSSD++ plots show the relationship between the exposure concentration and the predicted effect levels.
```{r}
# view pSSD plot for example:
pSSDs_water$`Tier3_Freshwater_Food Dilution`$pSSD_plot
```

```{r}
pSSDs_water$`Tier3_Marine_Tissue Translocation`$pSSD_plot
```
#### Arranged plot
A manuscript-ready pSSD++ with PNEC distribution plots for Tiers 3 and 4 are generated.
```{r}
pSSDs_water$`Tier3_Freshwater_Food Dilution`$arranged_plot
```

Sediment plots use the same access pattern, but with sediment environments:

```{r}
pSSDs_sediment$`Tier3_Freshwater Sediment_Food Dilution`$pSSD_plot
```

The cache and figure outputs default to temporary directories; override the paths above to persist artifacts between sessions.

### Combining environments
In some cases, one may want to combine pSSD outputs across different environments (e.g., freshwater + marine sediment due to low number of species and/or mixed species environment such as an estuary). This can be done accomplished by overwriting the environment variable in the toxicity dataset before running the pSSD workflow. An example is provided below.

```{r}
# overwrite environment variable
tox_data_combined <- tox_data |> dplyr::mutate(env_f = "Freshwater")

#MC-Align
MC_sim_df_combined <- MC_sim_align_parallel(
  tox_data = tox_data_combined, # toxicity data (ToMEx 2.0 is default)
  param_matrix = param_matrix, # parameter matrix generated in prior cell
  n_sim = 10, # number of MC alignment draws per species (must not exceed nrow(param_matrix))
  x1D_set = 1, # minimum particle size to generate distributions to (in microns)
  x2D_set = 5000, # maximum particle size to generate distributions to (in microns)
  num_cores = num_cores # number of cores to use for parallel processing (specify )
)

# Filter results to valid data
results_df_food_sediment_combined <- dplyr::filter(
  MC_sim_df_combined,
  ingestible != "not ingestible",
  particles_kg_food_dilution > 0,
  Group != "Algae") |>
  dplyr::mutate(dose_new_particles_kg = particles_kg_food_dilution) |>
  tidyr::drop_na(particles_kg_food_dilution)

# build ERM registry for pSSD (just one tier for food dilution for minimal reprex)
erm_registry_sediment_combined <- list(
  "Food Dilution" = list(base = results_df_food_sediment_combined, t3_t4 = results_df_food_sediment_combined),
  "Tissue Translocation" = list(base = NA, t3_t4 = NA)
)

# run pSSD workflow
pSSDs_sediment_combined <- make_all_pSSDs(
  MC_sim_df = MC_sim_df_combined, # ensure aligned data is included
  environments = c("Freshwater Sediment"), #marine and freshwater combined to fresdhwater
  erm_registry = erm_registry_sediment_combined,
  sim = 3, 
  base_cache_dir = file.path(tempdir(), "pssd_cache_combined"), # directory to store cached pSSD objects (important when running high number of simulations to resume if needed)
  base_output_path = file.path(tempdir(), "pssd_figures_combined"), # directory to store output figures
  overwrite_cache = TRUE # whether or not to overwrite cached pSSD objects
)

# inspect results
pSSDs_sediment_combined$`Tier3_Freshwater Sediment_Food Dilution`$pSSD_plot
```

---

## Glossary

Quick reference for terminology used throughout this vignette and the
accompanying manuscript (Coffin et al. 2026). See also `?PSSDplusplus_glossary`
for the full in-package reference.

### Statistical and modelling methods

| Term | Definition |
|------|-----------|
| **LHS** | Latin Hypercube Sampling — stratified-random sampling ensuring uniform coverage of parameter space |
| **MC (Monte Carlo)** | Repeated random sampling used to propagate parameter uncertainty through the full workflow |
| **Sobol' sequence** | Quasi-random low-discrepancy sequence underlying the LHS parameter matrix |
| **Power law (alpha)** | Exponent describing the size–abundance relationship of particles in a polydisperse environment |
| **Bootstrapping** | Resampling method for estimating SSD confidence intervals |
| **RCI** | Relative Confidence Interval: (95th − 5th percentile) / median |

### Ecological risk assessment

| Term | Definition |
|------|-----------|
| **SSD** | Species Sensitivity Distribution — model of variation in sensitivity across species |
| **PSSD** | Probabilistic SSD — preserves species-level variability via per-species distributions |
| **PSSD+** | Adds probabilistic assessment factors (Wigger et al. 2020) |
| **PSSD++** | This package — PSSD+ with MC-propagated ERM alignment uncertainty |
| **HC5 / HC10** | Hazard Concentration protecting 95% / 90% of species |
| **PNEC** | Predicted No Effect Concentration — the derived threshold |
| **Tier** | Data-filter level 1–4: protective (1–2, all endpoints) → predictive (3–4, organism/population) |
| **AF** | Assessment Factor — converts acute or ECx/LCx values to chronic NOEC equivalents |
| **cv_uf** | Coefficient of variation for the uncertainty factor distribution in PSSD++ |

### Microplastic-specific terms

| Term | Definition |
|------|-----------|
| **ERM** | Ecologically Relevant Metric — a particle- and species-specific measure of MP exposure effects |
| **Food dilution** | ERM: MPs dilute caloric intake; aligned to ingestible particle volume/L |
| **Tissue translocation** | ERM: MPs cross the gut epithelium; aligned to ingestible surface area/L |
| **Bioaccessibility** | Fraction of environmental MPs accessible to an organism based on size and physiology |
| **Translocation** | Movement of particles from gut to tissues, predicted by logistic regression on particle length |
| **ToMEx 2.0** | Toxicity of Microplastics Explorer v2.0 — the bundled ecotoxicity database |

### Effect metrics

| Acronym | Full name | Notes |
|---------|-----------|-------|
| NOEC | No Observed Effect Concentration | Target endpoint for all tiers |
| LOEC | Lowest Observed Effect Concentration | Converted to NOEC via `af.noec` |
| ECx | Effect Concentration at x% | x = percentage effect (e.g., EC50) |
| LCx | Lethal Concentration at x% | x = percentage mortality |
| HONEC | Highest Observed No Effect Concentration | Excluded — not rigorously determined |

### Key polymer abbreviations

PA = Polyamide; PE = Polyethylene; PET = Polyethylene Terephthalate;
PP = Polypropylene; PS = Polystyrene; PTFE = Polytetrafluoroethylene;
PUR = Polyurethane; PVC = Polyvinyl Chloride.
